Impairment of lipid homoeostasis causes accumulation of protein aggregates in the lysosome

Version 1

Abstract

Abstract

Protein aggregation increases during aging and is a pathological hallmark of many age-related diseases. Protein homeostasis (proteostasis) depends on a core network of factors directly influencing protein production, folding, trafficking, and degradation. Cellular proteostasis also depends on the overall composition of the proteome and numerous environmental variables. Modulating this cellular proteostasis state can influence the stability of multiple endogenous proteins, yet the factors contributing to this state remain incompletely characterized. Here, we perform genome-wide CRISPRi screens to elucidate the modulators of proteostasis state in mammalian cells using a fluorescent dye to monitor endogenous protein aggregation. These screens recovered components of the known proteostasis network, and uncovered a link between protein and lipid homeostasis. We subsequently showed that increased lipid uptake and/or disrupted lipid metabolism led to increased lysosomal protein aggregation and, concomitantly, accumulation of sphingolipids and cholesterol esters. Surprisingly, lysosomal proteostasis impairment by lipid dysregulation is independent of lipid peroxidation or changes in lysosomal stability, nor is it caused by effects on many other aspects of lysosomal or proteasomal function. These results suggest that lipid dysregulation may have primary effects on the stability of endogenous proteins, potentially through direct biophysical mechanisms.

Introduction

Protein aggregation is a hallmark of aging and age-related pathologies, including multiple neurodegenerative diseases and amyloidoses that affect individual or multiple organs 1,2. Dozens of proteins that are normally soluble may form aggregates that are traditionally associated with specific diseases -- e.g. tau tangles and amyloid plaques with Alzheimer’s Disease, and transthyretin deposits with transthyretin amyloidosis 2. Emerging evidence in the past decade or so, however, suggests a more complex relationship among different aggregation-prone proteins, either directly or through interaction with the cellular machinery regulating proteostasis. For instance, experiments in yeast, worms, and mammalian cell culture have demonstrated considerable cross-talk and cross-sensitization between classic aggregation-prone proteins 3. Further, many more non-disease associated proteins have been found to exhibit reduced solubility and lose their native structure during normal aging 4. Intriguingly, these insoluble non-disease associated proteins can seed the formation of classical amyloids in vitro, as well as contribute to functional decline in aging animals 5. Together, the evidence suggests that systemic changes in the state of cellular proteostasis may underlie proteinopathy, and highlights the importance of expanding the scope of studies beyond specific disease-associated protein species.

The core proteostasis network is well described and involves the synthesis, folding, trafficking and degradation of proteins 6–8. Most of these components closely follow the life cycle of proteins, including 1) pathways that modulate protein synthesis and degradation, such as mTORC, ribosome biogenesis and processing, UPS, and autophagy; 2) molecular chaperones, which aid protein folding and refolding; and 3) stress response pathways that regulate the activity of 1) and 2) in response to accumulation of unfolded or misfolded proteins, such as ERAD and UPR. Nevertheless, the components and network topology varies depending on cell type and cell state. Furthermore, mechanisms that underlie shifts in cellular proteostasis state, especially ones that do not directly involve core nodes of the proteostasis network, remain largely unknown.

Unbiased, whole-genome screening is a powerful tool for discovering cellular components and pathways that are involved in modifying a specific phenotype. Previous studies have applied genetic screens on different cell models to uncover, for specific proteins, modifiers of their aggregation, toxicity, or propensity for aggregation to propagate from seeds. These systems are often driven by overexpression of aggregation-prone proteins and/or extracellular seeding of pre-formed fibers 9–11. Thus, while much progress has been made in understanding the homeostasis of several discrete proteins, the mechanisms influencing the broader cellular proteostasis network remain largely unknown. In particular, we are interested in cellular mechanisms that affect the aggregation state of endogenous proteins, as these may be the most germane to age-associated loss of proteostasis.

To uncover novel components and pathways that regulate endogenous protein homeostasis, we applied CRISPR interference (CRISPRi) technology to perform unbiased, genome-wide screens on K562 cells using an aggresome-staining dye, ProteoStat. ProteoStat is known to label both experimentally induced aggresomes and aggregates that accumulate with age in the CNS of mammals 12–15. Our screens uncovered a link between lysosomal lipid homeostasis and proteostasis. Increased levels of sphingolipid and cholesterol esters, either as a result of increased lipid uptake or decreased lysosomal degradative capacity, lead to increased protein aggregation in cells. Lipid dysregulation induces lipid peroxidation, however this was not required for lipid-induced protein aggregation. Furthermore, lipid dysregulation induces protein aggregation without compromising lysosomal membrane integrity, lysosomal protease activity, or proteasome activity. Together, this work provides strong evidence that supports an emergent role of lipid dysregulation in disrupting endogenous protein homeostasis, independent of currently known mechanisms of action.

Results

FACS-based reporter for endogenous cellular proteostasis state

ProteoStat has been used to stain endogenous protein aggregation in fixed cells. Like Thioflavin T, ProteoStat fluorescence increases when rotationally confined by binding to cross-beta sheets in amyloidal structures. The dye has recently been used in studies to visualize aggregated proteins in mouse NSCs and in C. elegans 12,13,15,16. To verify that ProteoStat can be used to report changes in proteostasis in human cell cultures, we inhibited proteasomes in K562, U2OS, and A549 cells and stained them for microscopy or flow cytometry analysis. As expected, proteasome inhibition by MG132 increased ProteoStat intensity, with numerous bright puncta around the nucleus (Fig S1A). ProteoStat-high cells obtained by fluorescence-activated cell sorting (FACS) from cells treated with submaximal dose of MG132 likewise showed numerous bright puncta, whereas ProteoStat-low cells displayed diffuse staining of lower intensity, indicating that this tool can be used to report and isolate cells in different proteostasis states (Fig 1A).

CRISPRi screens with ProteoStat recovered major cellular proteostasis components

A) MG132-treated cells (at submaximal dose) were stained, sorted into top and bottom quartiles (ProteoStat-high and ProteoStat-low, respectively), and imaged. Red: ProteoStat; Blue: Hoechst 33342. B) FACS schematic. Stained cells are sorted using gates that correct for correlation between staining intensity and cell size. C) Volcano plot visualizing CRISPRi screen results (output from MAGeCK analysis with biological duplicate experiments). Genes in select GO Terms identified with GSEA (in D) are highlighted. D) GO Terms identified with GSEA. Positive Enrichment scores indicate an increased likelihood of genes in the GO Term to increase ProteoStat phenotype when knocked-down.

Genome-wide screen for modifiers of ProteoStat recovered major known components of the proteostasis machinery

To identify modifiers of endogenous proteostasis, we performed FACS-based, pooled screens in K562 CRISPRi cells, which stably express KRAB-dCas9. Briefly, cells carrying individual sgRNAs were fixed, stained with ProteoStat, and separated by FACS, using gates that correct for correlation between staining intensity and cell size, to obtain cells with ProteoStat intensity in the top or bottom quartiles (ProteoStat-high and ProteoStat-low, respectively; Fig 1B). Illumina sequencing of processed samples allowed us to quantify the frequency of each sgRNA in each of these two populations. This experiment was performed twice, and the resulting sgRNA frequencies were analyzed using the MAGeCK package to evaluate the effect of knockdown (KD) of each gene on the proteostasis state as reported by ProteoStat staining 17.

The screens identified 239 genes whose KDs increased ProteoStat staining, whereas 607 genes were enriched in the ProteoStat-low population at a FDR cutoff of 0.05 (Fig 1C). The results of a select subset of genes were validated by repeating the screen at a reduced scale (batch retest; Fig S1B). We performed gene set enrichment analysis (GSEA) to capture ontologies in which genes showed consistent and significant effects in the same direction. Many of these genes and pathways were previously known to be involved in cellular proteostasis (Fig 1D). These include ribosomal proteins in both the small 40S or the large 60S ribosomal subunits, ribosome biogenesis and processing, proteasome subunits, lysosomal proton subunits (vATPases), and certain chaperones. Knocking down subunits of the proteasome complex leads to increases in ProteoStat signal, consistent with pharmacological perturbation with MG132 in our validation experiments. Similarly, disrupting vATPases increases ProteoStat staining, likely via deacidification and impairment of lysosomes, an important center where multiple catabolic processes converge. These effects on ProteoStat signal following inhibition of the proteasome and lysosomal acidification were replicated using pharmacological inhibitors (Fig S1C). In contrast, knocking down many chaperones, including HSPA4, HSP90AB1, DNAJB6, and several subunits of TRiC chaperonins, paradoxically reduces the ProteoStat signal. Perturbing chaperones can trigger the newly described safeguard against protein aggregation mechanism that targets certain newly synthesized aggregation-prone proteins to the lysosome, notably before large visible aggregates are formed 18. Alternatively, the paradoxical effect of chaperone inhibition on ProteoStat signal could be due to upregulation of heat shock response as a negative feedback following the initial HSP70 or HSP90 inhibition 19–21. Indeed, pharmacological inhibition of chaperones led to increased ProteoStat signal after 30 minutes, but decreased signal after 24 hours (Fig S1C). Finally, we found that KD of many ribosomal proteins and ribosome assembly factors leads to a decrease in ProteoStat signal, consistent with reduced input of proteins into the proteostasis network 22.

Surprisingly, KD of most of the major components directly involved in macroautophagy did not lead to significant enrichment in either direction. These include the core autophagosome formation components (e.g. ATG5, ATG10, ATG12, and the ATG8 protein family) and the autophagy adaptors (NBR1, TAX1BP1, OPTN, SQSTM1 etc). Though the screens were performed under nutrient-replete conditions, CRISPRi perturbations of these genes in K562 under similar growth conditions have previously been demonstrated to influence autophagic flux 23. These results suggest that basal autophagy plays a negligible role in clearing endogenous amyloid-like structures in this system.

Lipid uptake and metabolism modulate proteostasis

The strong association of lysosomal vATPases with Proteostat phenotype is likely due to the need to maintain a lysosomal pH of ∼4.5-5.0 for optimal activity of the hydrolases that break down proteins, lipids, and nucleic acids. Failure to maintain this acidic pH can lead to lysosomal storage of undegraded macromolecules. This association prompted us to examine all lysosome-associated genes identified in the screen. We found that in addition to the vATPases, several genes that pleiotropically modulate the delivery or activity of multiple lysosomal enzymes simultaneously (e.g. GNPTAB, IGF2R, and PSAP), as well as multiple components involved in lipid uptake and metabolism (e.g. LDLR, MYLIP, and SCAP; Fig 2A) significantly affected the Proteostat phenotype. However, no perturbations of more discrete activities (e.g. individual cathepsins, lipid hydrolases or solute carriers) did so, suggesting that proteostasis is generally resilient to discrete lysosomal impairment.

Lipid uptake and metabolism modulate proteostasis

A) Volcano plot (same as Fig 1B) highlighting select genes involved in lipid uptake and metabolism. B & E) Comparing effects on ProteoStat phenotypes of single-gene vs double-gene KDs (n ≥ 2) C) Effects of candidate gene KDs on ProteoStat phenotypes in cells maintained in media with regular or delipidated FBS for 4 days. (n = 3) D) Effects of supplementing free cholesterol or human LDL in rescuing the ProteoStat-lowering phenotype of cells grown in media with delipidated FBS (DL media) for 5 days (n = 3). Phenotypes in B-E were normalized first to an internal staining control, and then to cells infected with non-targeting guides. (*: p<0.05; **: p<0.01; t-test)

KD of low density lipoprotein receptor (LDLR) reduced ProteoStat staining. Correspondingly, KD of the E3 ubiquitin-protein ligase MYLIP, which targets LDLR for sterol-regulated degradation, led to one of the strongest increases in ProteoStat staining. The effect of MYLIP KD on ProteoStat is via upregulated LDLR levels rather than accumulation of other MYLIP targets, as the effect of LDLR KD is epistatic to MYLIP in cells with double KD (Fig 2B). Cultured cells take up LDL - lipoproteins rich in cholesterol esters - through LDLR and the endo-lysosomal pathway. Once inside the acidic environment of the lysosomes, cholesterol ester is hydrolyzed into fatty acid and free cholesterol, the latter of which is trafficked to the ER. Excess cholesterol exerts negative feedback on LDLR levels via two mechanisms: 1) oxysterols, oxidized derivatives of cholesterol, activate the liver X receptor (LXR), in turn increasing MYLIP expression and LDLR degradation 24; 2) high cholesterol concentration in the ER sequesters the SCAP-SREBPs complex, suppressing its activity and in turn reducing LDLR expression as well as de novo cholesterol synthesis 25. A number of positive regulators of the SREBP pathways (SCAP and MBTPS2) also reduced ProteoStat staining when knocked down, though genes along the mevalonate -- de novo cholesterol synthesis -- pathway did not consistently score in our screen, indicating the difference between cholesterol that is taken up vs synthesized de novo.

Chronic depletion of exogenous LDL using media that contains delipidated serum reduces ProteoStat staining to a level similar to LDLR KD and is sufficient to abrogate the effect of MYLIP KD (Fig 2C), demonstrating that the effect of these genetic perturbations is mediated through lipid uptake and not direct degradative action of MYLIP on putative aggregation-prone substrates. Furthermore, supplementing exogenous cholesterol or human LDL either partially or completely reverses the effect of LDL depletion (Fig 2D). We also examined the possibility that ProteoStat may directly stain cholesterol by acute cholesterol extraction using methyl-β-cyclodextrin (MBCD). MBCD acutely removed most free cholesterol as determined by Filipin III staining, but did not affect ProteoStat signal (Fig S2A,B). In neither untreated nor MG132-treated cells did ProteoStat and Filipin III staining overlap, indicating that ProteoStat does not stain free cholesterol or any major components of the plasma membrane (Fig S2C). Furthermore, in an in vitro biphasic lipid bilayer system (giant unilamellar vesicles, GUVs), ProteoStat stained the lipid bilayer poorly and did not preferentially localize to the cholesterol-rich phase (Fig S2D). Together, the evidence suggests that increased flux of LDL uptake and/or the levels of lipid metabolites generated during the process lead to increased protein aggregation.

Our screen also implicated genes involved in lysosomal sphingolipid metabolism and lysosomal protein trafficking (Fig 2A). KD of GNPTAB, IGF2R, and prosaposin (PSAP) all led to increased ProteoStat staining. GNPTAB, a Golgi-resident enzyme, catalyzes the first step of the formation of mannose 6-phosphate (M6P) markers on lysosomal enzymes. The transport of these enzymes from the golgi to the lysosomes is facilitated by binding to M6P receptors, including IGF2R (also known as the cation-independent M6P receptor). PSAP is the precursor of saposin peptides, which are required to activate lysosomal sphingolipid catabolizing enzymes. The effect of PSAP KD was dramatically exacerbated by MYLIP KD and alleviated by LDLR KD (Fig 2E), indicating that lipid uptake can impinge on proteostasis through its interaction with lipid catabolism in the lysosome. The individual lysosomal hydrolases activated by saposins, such as GBA, GALC, and LIPA, did not rise above the detection threshold in our screen, again suggesting that proteostasis as measured by ProteoStat is generally resilient to discrete lysosomal impairment or changes in individual lipid species.

The evidence so far suggested a central role for lysosomal function in the ProteoStat phenotype under lipid-related perturbations, prompting us to further characterize ProteoStat staining, especially its subcellular localization, in reference to known biology of aggresomes. In one model of aggresome formation, misfolded and aggregated proteins are poly-ubiquitinated, transported along microtubules, and coalesced into perinuclear aggresomes, which are then detected by autophagy receptors to induce autophagosome formation, eventually delivering the cargo to the lysosomes for degradation 26. ProteoStat staining was initially reported to colocalize with mono-/poly-ubiquitinated proteins, p62, and LC3, and has been increasingly used as a general reporter for protein aggregation 12,13,15,16. To our surprise, we did not see colocalization of MG132-induced ProteoStat puncta with mono-/poly-ubiquinated proteins (Fig S2E). Instead, ProteoStat puncta colocalized with lysosomes (Fig S2E,F). Lysosome-accumulating ProteoStat-positive aggregates were reported in quiescent neural stem cells but not their activated counterparts 12, nor in other rapidly dividing cell types. Together, these results suggest that ProteoStat staining revealed an unexpected subset of lysosomal protein aggregates distinct from classical aggresomes, despite comparable regulation by known proteostasis components 12,13,15,16.

Sphingolipid and cholesterol ester mediate the effect of lipid perturbations on proteostasis

The links between proteostasis and lipid metabolism, especially in the lysosomes, provoked the question of whether these changes were mediated by a small number of lipid species, or if multiple lipid species were involved. To address this question, we performed lipidomics analysis on cells in which lipid metabolism was perturbed, either genetically by CRISPRi, or environmentally by lipid depletion. These perturbations were selected from the screen results to represent a range of effects on ProteoStat staining. When only a core set of perturbations, which have direct connections with lipid uptake and metabolism, were included, we found remarkable correlation between ProteoStat staining and individual levels of most lipids in the sphingomyelin and cholesterol ester classes (R2 of up to 0.968 and 0.843, for sphingomyelins and cholesterol esters, respectively, Fig 3A). Indeed, each of these two lipid classes -- the collection of sphingomyelins or cholesterol esters -- had significantly stronger correlation with ProteoStat phenotype when compared with the full collection of lipid compounds assayed (Fig 3B, see Fig S3A for the lipid classes ceramides (Cers), lactosylceramides (LacCers), or hexosylceramides (HexCers) as counterexamples). In contrast and for comparison, when targets that do not impinge directly on lipid metabolism (e.g. AMBRA1 or CHMP2A KD) or else expected to have strong pleiotropic effects beyond lipid metabolism (e.g. ATP6V1E1 KD) were considered, the correlation between ProteoStat staining and SM or CE lipid levels was substantially lower (insets in Fig 3A). These results further highlight the multimodality of proteostasis regulation, where lipid dysregulation represents one, among other, major components.

Sphingolipid and cholesterol ester mediate the effect of lipid perturbation on proteostasis

A) ProteoStat phenotype vs abundance of indicated lipid species in cells with CRISPRi KD of only a core set of lipid-related target genes, or including other ProteoStat perturbing genes (insets). Both ProteoStat phenotype and lipid abundance were normalized to cells carrying non-targeting guides. Least squares linear regression models (dotted red lines) were generated for the ProteoStat phenotypes and corresponding lipid abundances, and the Coefficient of determination, R2, values for each model is shown. B) Histograms of R2 values obtained for all lipid species within each indicated lipid class as in A. p-values were calculated using K-S test comparing R2 values within each lipid class versus that of all lipids. C) Schematic of the series of constructs expressing different variations of PSAP (full-length, truncated, carrying mutations deleterious to individual Sap peptides, or individual functional Sap peptides alone). Colored boxes depict individual Sap peptides. Deleterious mutations are depicted by X’s on the corresponding Sap peptide. C) ProteoStat phenotypes in cells overexpressing various PSAP variants in E, in conjunction with either PSAP KD or non-targeting control (n=3, #: p>0.05; t-test).

To determine if any lipid species alone contributes critically to protein aggregation induced by lipid dysregulation, we leveraged the modularity of the highly conserved PSAP. Upon delivery to the lysosome, the proprotein is cleaved proteolytically into four active peptides, saposins (Saps) A, B, C, and D. Saps are involved in the hydrolysis of glycosphingolipids (GSLs) as enzymatic activators, facilitating the interaction between soluble lysosomal hydrolases and GSLs embedded in the membrane 27,28. Though homologous, distinct residues between the saposins within the saposin-hydrolase interface confer hydrolase specificity on each saposin (e.g. SapA binds to Galactocerebrosidase, GALC, and stimulates galactosylceramide hydrolysis, Fig S3B) 29. Deficiency in full-length PSAP or individual Saps (by truncation, inactivating mutations, or reduced expression) lead to various lysosomal storage disorders, with neuropathology and often dysfunction of other organs. Depending on which Sap is deficient, distinct GSL species are found to accumulate in the storage cells. We therefore performed rescue experiments using different PSAP variants to perturb the level of individual GSLs more specifically 30. We created expression constructs for single Sap add-back as well as PSAPs that are full-length, truncated (Sap D deficient), or with various mutations that inactivate a single or all four Saps (Fig 3C, see Methods for a detailed description of variants). Full-length PSAP largely rescued the increase in ProteoStat level induced by PSAP KD (p=0.07 vs non-targeting control, no rescue), while the variant with all four Saps inactivated (PSAP ΔABCD) failed to rescue (p=0.74 vs PSAP KD, no rescue; Fig 3D). Furthermore, addition of any single Sap did not fully rescue the KD phenotype. Accordingly, inactivating any single Sap also did not fully abrogate PSAP’s ability to rescue. Together, the evidence suggests that proteostasis impairment from sphingolipid accumulation is pleiotropic and is likely attributable to the action of multiple but not a single lipid. This is consistent with the observation that, in our CRISPRi screens, none of the individual GSL hydrolases significantly increased ProteoStat level.

Lipid perturbation affects proteostasis without gross impact on lysosomal and proteasomal functions

Sphingomyelin and cholesterol metabolism were previously linked with several diseases with proteinopathy, including Alzheimer’s disease (AD) and the lysosomal storage disease Niemann Pick’s type C (NPC) 30–33, yet the link between these lipid species and proteostasis remains largely unclear. We hypothesized that lysosomal function may be impaired by accumulation of sphingomyelins and cholesterol esters, as 1) the lysosome plays an important role in metabolizing cholesterol esters and sphingolipids, 2) our screen implicated lysosomal acidification as an important modulator, and 3) lysosomal phenotypes are central to various lysosomal storage diseases. Finally, multiple lines of recent evidence suggest that sphingomyelin and/or cholesterol dysregulation may induce lipid peroxidation 34,35, lysosomal membrane permeabilization (LMP) 32,36,37, and impinge on the activity of the proteasome and lysosomal proteases 33,38. We therefore investigated different aspects of lysosomal health, including lysosomal number, size, protease activity, pH, and membrane integrity, as well as lipid peroxidation and proteasome activity, under several lipid perturbations.

To monitor lysosomal membrane integrity, we used a TagBFP-fused Galectin-3 (Gal3) reporter, which is normally diffuse in the cytoplasm, but binds to glycans on the lumenal side of lysosomal membranes that are exposed when the membrane is damaged. We imaged the reporter in cells under a combination of candidate gene KDs and/or acute LMP induced by L-leucyl-L-leucine (LLOMe), a lysosomotropic membranolytic agent (Fig 4A). CHMP2A KD, as a positive control, led to robust LMP in this assay. Unexpectedly, LDLR KD, which results in reduced CE and SM levels and reduced ProteoStat phenotype, moderately sensitized LLOMe-induced lysosomal damage. This result can be reconciled with the observation that physiological levels of lysosomal cholesterol may protect cells against lysosomal damage 39. In contrast, KDs of MYLIP, PSAP, and IGF2R did not affect baseline or LLOMe-induced lysosomal damage. Together, these results do not support increased LMP as an upstream determinant of protein aggregation (Fig 4B).

Effect of lipid perturbation on proteostasis is not mediated through the health and function of lysosomes and proteasomes

A) Representative fluorescence microscopy images of K562 cells expressing TagBFP-GAL3, either untreated or treated with LLOMe for 1.5 hours. B) Gal3 puncta per unit cell area quantified in fluorescence microscopy images taken with CRISPRi cells having various target gene KDs and either untreated or treated with LLOMe (n=2). C-F) Flow cytometry quantification of phenotypes in K562 cells with various target gene KDs. C: TR-X fluorescence (after protease-induced unquenching) as an indicator of lysosomal protease function; D: Ratio of lysosome-targeted sfGFP fluorescence (pH sensitive) to mCherry fluorescence (pH insensitive) as an indicator of lysosomal pH; E: LiperFluo staining as an indicator of lipid peroxidation level; F: LysoTracker-Red fluorescence as an indicator of lysosomal mass and pH. G-H) Efficacy of the 5 antioxidant cocktail (5AO) in rescuing the effects of target gene KDs on G) LiperFluo or H) ProteoStat phenotypes. (*: p<0.05; **: p<0.01; t-test) I) Flow cytometry-based assay with the Proteasome Activity Probe was validated with proteasome inhibitors. J) Effect of target gene KDs on proteasome activity.

To monitor lysosomal protease activity, we incubated cells with BSA conjugated to self-quenched red BODIPY TR-X (DQ-Red BSA). Uptake and delivery of these molecules to the lysosome lead to BSA hydrolysis and unquenching of the BODIPY fluorescence. As expected, impeding lysosomal acidification by ATP6V1E1 KD or lysosomal membrane integrity by CHMP2A KD led to reduced BODIPY unquenching. However, lipid-modifying perturbations affected lysosomal protease function only modestly and incoherently with respect to the ProteoStat phenotype, with MYLIP KD reducing BODIPY unquenching and PSAP KD trending in the opposite direction (Fig 4C).

Changes in lysosomal pH (pHLys) can be detected with high sensitivity using a ratiometric reporter (pHLARE), which consists of a pH-sensitive fluorescent protein (super-folder GFP, sfGFP) and a pH-insensitive one (mCherry) flanking rat LAMP1 40. Here, we stably expressed pHLARE, and validated the sensor for use with flow cytometry, using a buffer series with pH ranging from 4 to 7.5, in the presence of ionophores (Fig S4A). The sensor also correctly reported lysosomal deacidification as expected with ATP6V1E1 KD, and demonstrated that lysosomal membrane damage induced by either CHMP2A or CHMP6 KD led to modest increase in pHLys (Fig 4D). Nevertheless, lipid-modifying KDs -- MYLIP, PSAP, and LDLR -- did not significantly shift pHLys (Fig 4D), suggesting their effects on proteostasis are not mediated via lysosomal deacidification.

We investigated the level of lipid peroxidation and total lysosomal mass under lipid-modifying perturbations. We found that these perturbations modulate lipid peroxidation, as indicated by a fluorescent indicator, LiperFluo. Perturbations that increase or decrease sphingomyelin and cholesterol ester levels (MYLIP and PSAP KDs, vs LDLR KD) increase or decrease lipid peroxidation, respectively (Fig 4E). Interestingly, these lipid peroxidation phenotypes mirrored changes in LysoTracker intensity (Fig 4E-F, Fig S4B). As we did not observe these perturbations substantially altering pHLys (Fig 4D), the LysoTracker result suggested changes in lysosomal mass, potentially to compensate for lysosomal impairment that may or may not be a consequence of lipid peroxidation 32. To test if lipid peroxidation is necessary for the increased protein aggregation and lysosomal mass, we supplemented the media with five antioxidants (5AO), a cocktail commonly included in the media for culturing neurons --- a cell type that is particularly sensitive to oxidative stress. However, while this antioxidant cocktail was effective in reducing lipid peroxidation, it did not rescue the Proteostat or Lysotracker phenotypes (Fig 4G-H; Fig S4C). We therefore conclude that protein aggregation and lysosomal mass increase are likely upstream or independent of lipid peroxidation.

Finally, we investigated if the other major arm of protein catabolism -- proteasome -- is regulated by changes in lipid metabolism. A recent report suggested that high cholesterol ester levels in cells may hinder phospho-Tau degradation by proteasome inhibition 33. We labeled active proteasomes using a fluorescent probe, Me4BodipyFL-Ahx3Leu3VS (Me4BoVS), which covalently binds to active catalytic proteasome β-subunits in live cells, and quantified the labeling by flow cytometry (Fig 4I). HMGCR KD increased proteasome activity, consistent with the reported effect of statin treatment (Fig 4J) 33. Nevertheless, in conditions that strongly modulated sphingomyelin or cholesterol ester levels (KD of MYLIP, PSAP, or LDLR), we did not see significant increase or decrease of proteasome activity. In contrast, KD of IDI1, an enzyme downstream of HMGCR along the mevalonate pathway (but not FDFT1), strongly increased proteasome activity. We conclude that the proteasome modulating effect of statin likely arises from changes in other metabolites along the mevalonate pathway.

Discussion

Aging is associated with decreases in cellular proteostasis capacity and the accumulation of protein aggregates in diverse tissues. Despite extensive studies, factors modifying cellular proteostasis state, which has the potential to affect the state of many endogenous proteins, remain incompletely characterized. In this study, we performed genome-wide CRISPRi screens using a fluorescent dye sensitive to endogenous protein aggregation and uncovered links between lipid homeostasis and proteostasis. We discovered and characterized how changes in the uptake and metabolism of lipids, particularly sphingomyelins and cholesterol esters, lead to accumulation or depletion of these lipids and alter homeostasis of endogenous proteins.

Sphingolipid and cholesterol dysregulation has been implicated in various age-associated diseases. Intracellular sphingolipid accumulation has long been associated with lysosomal storage diseases, which share many features with age-related neurodegenerative diseases both at the level of physiology and cellular mechanism 41. Elevated extracellular LDL cholesterol is associated with atherosclerotic cardiovascular disease 42, where modifications, including oxidation, and misfolding of apoB-100 can trigger the formation of Thioflavin T-positive fibrils of aggregated LDL and stimulate uptake by macrophages and catabolism within their lysosomes 43,44. Consistent with our observations, MYLIP deficiency in mice leads to upregulated LDLR expression and increased cellular uptake of lipoprotein ApoE in brain tissues, but also increased uptake and clearance of extracellular aggregated Aβ peptides 45.These data affirm biological connections between lipoproteins and amyloidogenesis, the pathobiological effects of which depend on the localization of the accumulating lipids.

Prompted by the identification of lipid- and lysosome-modifying genes in altering proteostasis, we set out to understand the underlying molecular mechanism. Previously, accumulation of sphingolipid or cholesterol due to direct enzyme deficiency or impaired lipid transport was linked to reduced cellular proteostasis via impairment of lysosomal function, manifested variously as reduced lysosomal proteolytic capacity, LMP, lipid peroxidation, and aberrant mTORC1 signaling 32,38,46,47. Surprisingly, we showed in this work that many forms of lysosomal impairment are not necessary for lipid perturbations to impinge on proteostasis. More recently, Tian and colleagues discovered in iPSC-induced neurons that PSAP KD severely impairs GSL catabolism, leading to the accumulation of these lipids and, in turn, lipofuscin formation, lipid peroxidation, and eventual cell death by ferroptosis 34. Lipofuscins are lysosomal deposits of heterogeneous nature, consisting mainly of oxidized proteins, lipids, metal ions, and sugar residues 48. These deposits are largely seen in post-mitotic and senescent cells, and their accumulation is observed in aged tissues but also associated with many age-related degenerative diseases 48. Unsurprisingly, we did not observe lipofuscin in K562 cells with PSAP KD (Fig S4D), likely due to continuous dilution of damaged biomolecules by rapid cell divisions before their levels can reach a threshold to form crosslinked and undegradable granules. Despite no visible accumulation of lipofuscin, we nonetheless observed an increase in lipid peroxidation and concomitant increase in lysosomal mass, similarly to what’s reported in iPSC-derived neurons 34, suggesting that granular lipofuscin formation may be dispensable for ROS generation. Furthermore, we showed that lipid peroxidation is not required for lysosomal protein aggregation induced by lipid dysregulation.

We have ruled out requirements for generalized forms of lysosomal impairment in mediating protein aggregation in our system, suggesting that aggregation may be more directly coupled to the accumulation of sphingolipids and cholesterol esters. Sterol-rich lipoproteins can promote aggregation, as in the case of LDL(-) 43, and sterols themselves can mediate solubilization of inclusion bodies in yeast 49 and lanosterol can disaggregate cataractous crystallin 50,51. Alternatively, these lipid changes may indirectly influence the lysosomal milieu interieur by modulating ion channel activity 52 and changing the folding environment or altering the lysosomal proteome through changes in endolysosomal trafficking 53. We anticipate that future elucidation of these mechanisms will contribute to our molecular understanding of protein aggregation during aging and its impact on human health.

Materials and Methods

Cell line construction and screen procedure

CRISPRi cell line was obtained by transducing K562 cells with a lentiviral vector expressing the nuclease-deactivated Cas9 fused with transcriptional repressor KRAB (UCOE-SFFV-KRAB-dCas9), followed by antibiotic selection. Monoclonal cell lines were obtained by limiting dilution, and screened functionally for effectiveness at silencing test genes when treated with sgRNAs. Cells were maintained in suspension at a density between 125K to 1M cells/mL in RPMI complete media (with 10% FBS and 1x GlutaMax). 500 million stable CRISPRi cells were transduced with a sgRNA lentivirus pool (hCRISPRi_v2.1_top5 library, which includes 102,640 gene-targeting and 4,590 non-targeting sgRNAs) at an MOI of 0.3 for single integrants. After 6 days in culture, including antibiotic selection for cells carrying the sgRNAs with 2ug/mL puromycin, these cells were stained with fixable blue dead cell stain (Molecular Probes), then fixed, permeabilized and stained with ProteoStat Aggresome detection kit (Enzo) according to manufacturer’s protocol. Stained cells were sorted by FACS (BD FACSAria) into top and bottom quartiles (ProteoStat-high and ProteoStat-low, respectively), using gates that excluded dead cells and corrected for correlation between cell size and staining intensity. Sorted cell populations were pelleted, from which genomic DNA was extracted with phenol:chloroform:isoamyl alcohol and ethanol precipitation. sgRNA-containing sequences were amplified from the genomic DNA with two rounds of PCR, introducing multiplexing barcodes and sequencing adaptors in the process. Amplicons were then purified and sequenced with the HiSeq4000 platform (Illumina). Sequencing data were analyzed using the MAGeCK pipeline 17. GSEA was performed using the GSEAPY package 54,55.

Validation of hit genes by batch-retesting a reduced pool of CRISPRi guides

A subset of sgRNAs were selected and cloned into a reduced-scale library that consisted of 572 sgRNAs targeting 286 genes (87 enriched in ProteoStat-high, 99 enriched in ProteoStat-low, and 100 unenriched but abundantly expressed in K562 cells) and 300 non-targeting sgRNAs. sgRNA transduction in K562 CRISPRi cells, ProteoStat staining, FACS, library preparation and sequencing were performed similarly to the genome-wide experiment, except at a reduced scale (20 million cells transduced with sgRNA lentivirus pool, and >2.8 million cells each of ProteoStat-high and ProteoStat-low populations were obtained by FACS).

ProteoStat assays on cells with individual perturbations

K562 cells in multi-well plates were treated with individual perturbations and harvested for ProteoStat staining. To control for cell density-dependent variation in staining, harvested cells were spiked with untreated cells carrying appropriate fluorescent markers (e.g. TagBFP, mCerulean, or unlabeled, to be distinguished from treated cells) before fixation, permeabilization, and staining. ProteoStat fluorescence was quantified by flow cytometry (Fortessa, BD) with excitation at 488 nm and bandpass emission at 695/40 nm. ProteoStat intensity for each sample was first normalized to the internal spiked in (untreated) control, then to the corresponding negative control sample in the experiment (e.g. Non-targeting sgRNA). Individual perturbations were achieved by CRISPRi (sgRNA lentiviruses) or pharmacological inhibitors -- MG132 (EMD Millipore; 474790), Bafilomycin A1 (Invivogen; TLRL-BAF1), VER155008 (Selleckchem; S7751), Ganetespib (Selleckchem; S1159), 17-DMAG (Invivogen; ant-dgl), MBCD (Sigma; C4555), Bortezomib (Selleckchem; S1013), LLOMe (MedKoo; 597431), and Cumene Hydroperoxide (Sigma-Aldrich; 247502). Lipid deprivation was performed by substituting regular FBS with delipidated FBS (Lipoprotein depleted FBS, Kalen Biomed; 880100).

Immunofluorescence and cell staining by fluorescent probes

Cholesterol quantification was performed by formaldehyde fixing and staining cells with Filipin III (Abcam; ab133116) according to manufacturer’s instructions.

Immunofluorescence assays were performed with standard protocol. Briefly, cells were washed with PBS and fixed with 4% formaldehyde for 30 minutes at room temperature, then washed with PBS. Cells were permeabilized in blocking buffer (PBS +0.5% Triton-X 100 +5% goat serum) for 1 hour at room temperature, followed by incubating with blocking buffer + primary antibody at 4ºC overnight. Cells were then washed 3 times with PBS-T (PBS + 0.05% v/v tween20) at room temperature and incubated with blocking buffer + secondary antibody at room temperature for 1 hour. Cells were washed twice more with PBS-T and once with PBS before imaging. If co-staining with ProteoStat, the dye was included at 1:10000 during both primary and secondary antibody incubation. List of antibodies used: mouse-anti-LAMP1 (Abcam; ab25630), mouse-anti-mono/polyubiquitinated proteins (Enzo; BML-PW8810-0100), and Alexa Fluor Plus 647-conjugated goat-anti-mouse (Invitrogen; A32728).

Lipid peroxidation was quantified by incubating cells with Liperfluo (Dojindo Molecular Technologies; L24810) at 5uM for 1 hour at 37ºC. Cells were resuspended in cold PBS before analysis with flow cytometry (Fortessa, BD). Antioxidants, when included, consist of the following components at the corresponding final concentration: 5 ug/mL catalase (Sigma-Aldrich; C40), 2 ug/mL glutathione (Sigma-Aldrich; G6013), 15 U/mL superoxide dismutase (Sigma-Aldrich; S5395), 2 ug/mL DL-alpha tocopherol (Sigma-Aldrich; T3251), and 2 ug/mL DL-alpha tocopherol acetate (Sigma-Aldrich; T3001).

Lysosomal protease activity was monitored with DQ-Red BSA (Invitrogen; D12051). K562 CRISPRi cells carrying sgRNAs targeting corresponding genes (7 days after sgRNA introduction) were pelleted and resuspended in prewarmed trafficking media (RPMI + 1% serum + 1% GlutaMax + 10 nM HEPES) containing 10 ug/mL DQ-Red BSA and returned to incubator at 37ºC. After incubation for 1 hour, cells were washed once with cold PBS, resuspended in cold PBS, and analyzed by flow cytometry (Fortessa, BD) with excitation at 561 nm and bandpass emission at 610/20 nm.

Lysosomal mass was estimated by incubating cells under various genetic perturbations (7 days after sgRNA transduction) with LysoTracker Red DND-99 at 100 nM (Invitrogen; L7528). After incubation at 37ºC for 1 hour, cells were washed once with cold PBS, resuspended in cold PBS, and analyzed by flow cytometry (Fortessa, BD) with excitation at 561 nm and bandpass emission at 610/20 nm.

Proteasome activity was assessed similarly by incubating cells with genetic (7 days) or chemical (1 hour) perturbations with Me4BoVS (R&D system; I-190-050). After incubation at 37ºC for 2 hours, cells were washed once with cold PBS, resuspended in cold PBS, and analyzed by flow cytometry (Fortessa, BD) with excitation at 488 nm and bandpass emission at 525/50 nm.

Lipidomics experiment and data analysis

K562 cells under various CRISPRi or lipid perturbations were maintained, in triplicate, for 7 days. Cells were pelleted, resuspended in fresh medium, and returned to culture on Day 6 -- one day before up to ∼ 1 million cells per sample were harvested. Briefly, cells were pelleted and washed twice with PBS at 4ºC, then lysed using 800 uL of ice-cold MeOH:H2O (1:1,v/v) containing 3% by volume of lipid standards - Avanti 330707 SPLASH® LIPIDOMIX® Mass Spec Standard) at −20ºC, and transferred to a glass vial. Following the methanol-tert-butyl methyl ether Matyash extraction method 56, 800 uL of 100% MTBE was added into each sample, vortexed, and incubated on ice for 10 minutes to form phase separation. Next, after centrifuging samples at 3500RPM at 4°C for 5 minutes, 500 uL of the upper organic phase was collected into a clean glass vial, and the lower phase was re-extracted by the addition of 600 uL of 100% MTBE as described above. The organic phases were combined, dried under nitrogen, and resuspended in 150 uL of HPLC graded-ButOH/MeOH/H2O (2:2:1,v/v/v) with internal standards including: 5 ug/mL PE(18:0/18:0)-D70, PA(14:0/14:0)-D54, and 25 ug/mL LPC (16:0)-D31. Samples were analyzed using Thermo Scientific Vanquish UHPLC coupled with Q Exactive Plus Mass Spectrometer with 7.5 uL injection volume.

Analysis was done in both negative and positive mode ionization with separation on a Thermo Accucore C30 (250 × 2.1 mm, 2.6 μm polymer particles). Mobile phase A was 20 mM ammonium formate in MeCN:water (60:40,v/v); mobile phase B was 20 mM ammonium formate in isopropanol:MeCN (90:10, v/v) and a flow rate of 0.2 mL/min was used. The column was equilibrated for 7 min in 30% B prior to injection; followed by a gradient of 30-43%B from 0 to 7 min, 43-65% B from 7 to 12 min, 65-70% B from 12 to 30 min, 88% B at 31 min, 88-95% B at 51 min, 100% B at 53 min and 55 min, and 30%B from 55 to 60 min. The source parameters were as follows: Sheath gas: 40; Aux gas:15; Sweep gas: 1; spray voltage 3.1 kV, the capillary temperature of 275°C, and S-lens RF level of 50. The mass spectrometer was operated in data-dependent top-8 MS2 mode, with 14000 resolution setting and AGC of 3e6 for MS1, and 17,500 resolution setting and AGC target of 3e6 for MS2. Stepped, normalized collision energies of 20, 30, 40 were used.

Data were analyzed and visualized using the open source software MAVEN peak analysis program (https://github.com/eugenemel/maven) 57. Compounds were identified by comparison against an in silico generated lipid library of fragmentation patterns.

PSAP variant series and rescue experiments

A series of constructs expressing various PSAP variants were built into lentiviral vectors under a UBC promoter. The canonical sequence mRNA variant 1 (NM_002778.4) was designated as wild-type. The full-length, wild-type sequence was modified using the following known disease-associated mutations to generate deficient variants: SapA-deficiency (p.V70del), SapB-deficiency (p.N215H_C241S), SapC-deficiency (p.L349P_C388F). The SapD-deficiency variant was created by truncating the entire C-terminal starting at the first SapD residue (p.405_524del).

PSAP KD rescue experiments were performed by co-transducing K562 CRISPRi cells with lentiviruses expressing either a PSAP-targeting or non-targeting sgRNA, and in parallel one of the various PSAP variants. Transduced cells were selected with 2 ug/mL puromycin and 500 ug/mL geneticin for doubly transduced cells. Cells were harvested 6 days after transduction, spiked with untreated cells as internal control, then stained with ProteoStat and analyzed with flow cytometry as in other experiments.

Lysosomal membrane permeability assay

Stable LMP reporter cells were generated by transducing K562 CRISPRi cells with a lentiviral vector expressing TagBFP-Gal3, followed by selection with 10 ug/mL blasticidin. To measure the effect of gene KDs on LMP, these reporter cells were transduced with corresponding sgRNA-expressing lentiviruses and selected with 2 ug/mL puromycin. 5 days after sgRNA transduction, cells were plated on poly-D-lysine-coated CellCarrier Ultra 96-well plates (Perkin Elmer) at 12K cells in 200 uL regular media per well. 2 days after plating, cells were treated with 60 uM LLOMe (or left untreated) and imaged on Opera Phenix imaging platform (Perkin Elmer) at 15-minute intervals for 4 hours, with excitation laser at 375 nm and bandpass emission filter at 435-550 nm. Images were analyzed with Harmony software (Perkin Elmer) by cell segmentation and spot detection. The number of Gal3 puncta in each field of view (6 fields per condition) is normalized by the total segmented cell area and scaled to match the average cell area, to correct for cell density and averaged across replicates. The maximal response for each sample over the 4-hour time course was reported.

pHLys quantification by pHLARE

Stable pHLys reporter cells were generated by transducing K562 CRISPRi cells with a lentiviral vector expressing pHLARE (a fusion protein consisting of prolactin signal sequence, superfolder GFP [sfGFP], rat LAMP1, and mCherry developed by Webb et al.)40, and selected with 10 ug/mL blasticidin.

For validation, reporter cells were pelleted at 500 g for 5 minutes and resuspended in a buffer series at pHs ranging from 4 to 7.5 (140 mM KCl, 1 mM MgCl2, 1 mM CaCl2, 5 mM glucose and 10% citrate-phosphate-borate buffer at the target pH) 58 in the presence of ionophores (10 uM nigericin, 2 uM monensin). Non-reporter cells were treated in parallel to correct for changes in background autofluorescence under various perturbations. After incubation at 37ºC for 10 minutes, cells were analyzed with flow cytometry (Fortessa, BD), using excitation-bandpass emission settings of 488 nm-525/50 nm and 561 nm-610/20 nm for sfGFP and mCherry, respectively. To elucidate the effect of gene KDs on pHLys, CRISPRi cells with or without stable pHLys reporter were transduced with sgRNA-expressing lentiviruses, selected with 2 ug/mL puromycin, and analyzed similarly with flow cytometry, 7 days after sgRNA transduction.

Cytometry data were analyzed in the FlowJo software (BD). Median background fluorescence in each channel in the cognate non-reporter controls was subtracted from the sfGFP and mCherry fluorescence intensities in the pHLys reporter samples, and the ratio of adjusted sfGFP intensity vs adjusted mCherry intensity was reported.

Giant unilamellar vesicles (GUVs) and ProteoStat staining

GUVs with binary mixtures of saturated and unsaturated lipids were obtained by electroformation. Briefly, DOPC, DPPC, and free cholesterol were mixed in equal molar parts to a final concentration of 1mg/mL. TopFluor Cholesterol (Avanti Polar Lipids Inc.) was included at 1% of total lipids. The lipid mix was spread onto ITO-coated glass coverslips and dried overnight under vacuum. Dried lipids were hydrated with 300-400 ul of a 100 mM sucrose solution while applying a sinusoidal voltage (1.2 Vp-p and 10Hz) for one hour at 60ºC. GUVs were harvested, diluted 20-fold in 200 uL of a 100 mM glucose solution supplemented with either no ProteoStat, or ProteoStat at a final dilution of 10,000x and 500x and imaged at 30ºC in a spinning disk confocal fluorescence microscope (Nikon), using excitation/emission at 488 nm/525±18 nm (TopFluor) or 561 nm/605±35 nm (ProteoStat). Multi-channel single-plane images were acquired using NIS-Elements software and analyzed with ImageJ/Fiji 59.

Transmission Electron Microscopy

K562 CRISPRi cells carrying either non-targeting or PSAP-targeting sgRNA (6 days after sgRNA transduction) were fixed by directly mixing cells in suspension 1:1 with prewarmed 1x fixative (2% glutaraldehyde and 4% paraformaldehyde in 0.1 M sodium cacodylate buffer, pH7.4), followed immediately by gentle pelleting at 100 g for 10 minutes. Cell pellets were resuspended gently with fresh 1x fixative, and incubated at room temperature for 1 hour, followed by washing once with 2 mL 0.1 M sodium cacodylate buffer at room temperature and then transferring to 0.5 mL 0.1 M sodium cacodylate buffer, each time pelleting at 100 g for 10 minutes.

Fixed cells were stored at 4ºC until they were enrobed in gelatin and post-fixed with 1% osmium tetroxide, washed with ddH2O and stained with 1% uranyl acetate. Stained samples were dehydrated in a graded series of ethanol, infiltrated and embedded in epon resin. Samples were then trimmed and sectioned with an ultramicrotome and imaged on a JEM-1400 transmission electron microscope (JEOL).

Supporting information

eLife Assessment

Protein and lipid homeostasis are important for maintaining cellular functions but their crosstalk remains largely unknown. This important manuscript deals with this interesting topic and applies the powerful unbiased tools of somatic cell genetics to discover evidence suggesting a link between sphingolipids/cholesterol ester metabolism and lysosomal protein aggregation. However, in our view the experimental evidence in support of the paper's stated conclusions remains incomplete and measures that (i) go beyond reliance on the ProteoStat stain as the sole metric for aggregation and proteostatic challenge, (ii) biochemically back up the results and provide plausible mechanistic insights into the connection of lysosomal lipid handling and protein aggregation are needed.

Reviewer #1 (Public Review):

In this manuscript, Yong and colleagues link perturbations in lysosomal lipid metabolism with the generation of protein aggregates resulting from proteosome inhibition. The main tool used is the ProteoStat stain to assess protein aggregate burden in native cells (i.e. cells under no exogenous or endogenous stress). They initially use CRISPR-based genome-wide screens to identify several genes that affect this aggregate burden. Interestingly, knockdown of genes involved in lysosomal acidification was a major signature which led to identification of other culprit lysosome-associated genes that included ones involved in lipid metabolism. Subsequent CRISPR screen focused on lipidomic analysis led to identification of sphingolipid and cholesterol esters as lipid classes with effects on proteostasis. Despite using various tools of lysosomal function, acidity, permeability, etc, the authors couldn't identify the link between lysosomal lipid metabolism and protein aggregate formation. Nevertheless, the interrelationship of these two processes was the overall conclusion of this manuscript.

Although this work is interesting and thought-provoking, their approach to identify novel pathways involved in proteostasis is limited and this weakens the contribution of the paper in its current form.

Reviewer #2 (Public Review):

It is certainly an interesting observation that lipid homeostasis influences proteostasis, although this need not be considered so surprising given that many fundamental cellular processes are interconnected. The paper is deserves to be read, but the level of general interest would be greatly enhanced if the authors were able to take the story further mechanistically. This might be too much of an ask, but they should go further in excluding one very attractive alternative model: effects on proteasome activity. This explanation should be addressed definitively because the transcription factor that regulates proteasome subunit gene expression (Nrf1/NFE2L1) is processed in the ER and is therefore well placed to be influenced by membrane conditions, and because it is shown here that proteasome inhibition increase ProteoStat puncta. Indeed, some years ago it was published that Nrf1/NFE2L1 is inhibited within the ER membrane by cholesterol, and a more recent paper showed that in C. elegans it is activated by oleic acid through effects on ER membrane homeostasis and lipid droplet formation. The authors address proteasome activity only by using a dye that is not referenced. Here a much more solid answer is needed. In general, most conclusions in the paper rely essentially solely on ProteoStat assays. The entire study would be greatly strengthened if the authors incorporated biochemical or other modalities to substantiate their results.

The presentation would be improved greatly if the authors provided diagrams illustrating the pathways implicated in their results, as well as their models. As it is the paper falls flat at the end of the results in the absence of a mechanism to explain their findings. Diagrams would be helpful for focusing the reader on what IS learned from the work, which is important.

Author response:

We are submitting a revised manuscript with major additions that address the main concerns in the initial reviews. At the highest level, this revision provides i) orthogonal biochemical measurements that yield concrete evidence of lysosomal protein aggregates, and ii) a plausible mechanism linking lysosomal lipid handling and protein aggregation through disruption of ESCRT function. We believe these additions significantly improve the completeness of this study and the conclusions that can be drawn from the data.

Below are more specific highlights on the addition in this revision:

- We included orthogonal techniques (thioflavin-T staining and Lyso-IP followed by differential extraction) and confirmed the accumulation of RIPA-insoluble protein aggregates at the lysosomes in cells under lipid perturbation (Figure 3).

- We performed TMT-Proteomics and identified accumulation of insoluble ESCRT components at the lysosomes under lipid perturbation (Figure 4). Two new authors involved in this effort are added onto the manuscript.

- The ESCRT result prompted us to revisit lysosomal membrane integrity. With improved imaging conditions and analysis we were able to see increased membrane permeabilization under lipid perturbation. VPS4A overexpression partially rescued this phenotype, suggesting that lipid accumulation impairs ESCRT disassembly (Figure 5).

- Together, the results suggest that lipid perturbation impairs ESCRT function, compromising both lysosomal membrane repair and microautophagy, resulting in the accumulation of endogenous protein aggregates at the lysosomes (Graphical Abstract).

Reviewer #1 (Recommendations For The Authors):

(1) Perhaps the most prominent limitation of this work is the unilateral focus on native cells (i.e. cells under no endogenous or exogenous stress) as the model for protein aggregate formation. Furthermore, although the ProteoStat stain has been utilized by many investigators before, the sole reliance on this stain as the read-out for their assays is concerning. To compound the concern, the ProteoStat-positive puncta co-localize with lysosmal markers which was surprising even to the authors. All in all, it behooves the authors to test proteostasis in multiple parallel ways to actually define what they are studying. How is it possible that protein aggregates under native conditions are only co-localized with lysosomes? Are we really studying protein aggregates which should predominantly be cytoplasmic insoluble aggregates?

(a) They need to get away from a simple stain like ProteoStat and conduct co-stainings with other markers such as poly-ubiquitin antibodies and other chaperones to define what and where else exactly are these aggregates.

Co-staining with poly-ubiquitin was included in the original manuscript. We added orthogonal staining with another widely used amyloid dye, Thioflavin-T, and provided fine-grained quantification of lysosomal vs cytosolic localization of various signals (Figures S4A-C & 3A-B).

(b) They need to do Immunoblots with and without triton insolubility to see if these aggregates are insoluble as most would predict. They can do lysosomal isolation vs cytoplasmic to see if the insoluble aggregates are really lysosomal.

We performed Lyso-IP followed by differential detergent extraction to confirm the accumulation of insoluble proteins at the lysosomes (Figure 3C). Proteomic analysis identified some of these insoluble proteins as ESCRT subunits (Figure 4).

(c) They should compare aggregate formation in the native state versus cells with lysosomal inhibition via Bafilomycin or chloroquine versus cells with proteosomal inhibition. The lysosomal inhibition experiments are particularly informative given the lysosomal relevance they have uncovered.

We included other small molecule inhibitors and at different time points to compare the effect of different modes of proteostasis challenge (Figure S4A-D). Together with the ESCRT finding, our results suggest the role of microautophagy in our system, and provide a model of how ProteoStat- and/or ubiquitin- positive substrates become partitioned between the cytoplasm and lysosomes under different perturbations.

(d) Many protein aggregates which are too bulky for proteosome degradation will traditionally be dealt with by aggrephagy. Why is this not observed?

Knockdown of core macroautophagy components did not impact Proteostat intensity in our CRISPRi screen, suggesting that basal macroautophagy plays a negligible role in clearing endogenous amyloid-like structures in our experimental system. We provide an alternative model that these aggregates instead arrive at the lysosomes via microautophagy.

(2) After addressing #1, they can validate if the genes they identified by CRISPR screens are also important in modulation of protein aggregate burden in other systems. For example, if they inhibit lysosomes by Bafilo or Chloroquine to obtain protein aggregates and then Knockdown the identified genes in the CRISPR screens, will they get the same results?

We addressed the effect of different modes of proteostasis challenge as recommended above. Deacidifying the lysosomes alone causes intense protein aggregation (Figure S4A-D) and eventually cell death, and was thus not combined with other perturbations.

(3) They identify lysosomal lipid metabolism genes/pathways as the culprit for inducing proteostasis. In particular sphingolipid and cholesteryl ester species appear to be operational here. However, there are no specific lipids species or specific lipid metabolism gene that is causative. Rather, you have to knockdown entire processes to have an effect. This suggests that the focus on lysosome health (i.e. permeability, proteolysis, etc) is rudimentary. When you have to knockdown entire classes of lipids, this would indicate more broad effects on cellular lipids (including membrane lipids beyond the lysosome) and related cellular health?

We included data on the effect of knocking down MYLIP, PSAP, and as a comparison PSMD2 on the growth rate of K562 cells (Figure S5A). MYLIP and PSAP KDs, which cause predominantly an accumulation of lipids, do not impede cell growth. Increasing lipid uptake by MYLIP KD increases cell proliferation under our culture conditions, suggesting a general negative impact on cell health was not required for the association between lipid levels and protein aggregates.

(a) They conduct a superficial methyl-beta-cyclodextrin experiment with equivocal results. The use of MBCD for different time-courses to deplete various membrane cholesterol pools including the plasma membrane pool is important to ascertain what aspect of the cellular cholesterol is affecting proteostasis. MBCD +/- cholesterol reintroduction time-courses for rescue will also be key to determine the culprit cellular cholesterol pool.

The MBCD / Filipin experiment helped us determine that ProteoStat doesn’t directly stain cholesterol, nor any major plasma membrane components. Free cholesterol was implicated in neither the screen nor the lipidomics and was not the subject of targeted experiments.

(b) The same concept can be applied to sphingolipids. There are sphingolipids in abundance in multiple membrane compartments. Which ones are causal here? More nuanced evaluation of this with sphingolipid staining/tracking can be conducted.

We attempted experiments where sphingolipids were added back to cells grown in FBS-depleted media. Nevertheless, we were not able to consistently deliver these lipid species and doing so while ensuring the correct subcellular localization at physiologically relevant level would require substantial methods development.

(c) As part of this, are lipid rafts and/or caveolae being affected by the perturbations in cholesterol and sphingolipids? Lipid rafts are highly enriched in these 2 lipids which could link to their preteostasis observation.

Indeed, ceramides released from SM hydrolysis are proposed to self-assembled into microdomains with negative curvature that can promote the formation of intralumenal vesicles (Alonso and Goni, 2018; Niekamp et al 2022). We propose that SM accumulation may hinder this process by counteracting the negative membrane curvature and impede microautophagy.

(d) How about ER membrane lipids? The UPR and subsequent effects on proteostasis are intricately involved with ER lipid bilayer composition.

We did not perform lipidomics on ER membranes in this study, though we note that at steady state, sphingolipids and cholesterol esters are not expected to be enriched at the ER (Ikonen and Zhou, 2021). We checked whether lipid-related genetic perturbations induced the UPR in published perturb-seq data in K562 cells. Neither MYLIP nor PSAP knockdown induced a UPR.

In conclusion, the manuscript is interesting but the excitement over a link between lysosome-related lipid metabolism and proteostasis needs to be tamped until a more robust experimental approach is employed to generate supportive and corroborating results.

Reviewer #2 (Recommendations For The Authors):

- The paper has a number of grammatically awkward sentences. Editing these would enhance clarity.

- It is important to show the co-localization of aggregates with the lysosome. This is shown in supplements but should be in a main figure. Here the authors cite previous work indicating that ProteoStat puncta co-localize with ubiquitinated proteins and state that they do not see this, then essentially just move on. Is there an explanation for this discrepancy and can it be resolved? What do they think is really going on? What happens to levels of ubiquitinated proteins when lipid metabolism is perturbed as in these experiments?

We have included the lipid-induced lysosomal protein aggregation data in the main text (Figure 3A-B), and provided fine-grained quantification of the cytosolic-vs-lysosomal ProteoStat / Ub / ThT signals under different aggregate-inducing conditions (Figure S4A-D). We discuss these results in the main text and propose a model involving ESCRT-mediated microautophagy in the main text. This is supported further by the LysoIP-proteomics and LMP analysis.

- Please add an indicator of amino acid numbers to Fig. 3C.

These annotations are now included (now Figure S3C).

- The legend for 3D is mislabelled.

We have corrected the legend (now Figure S3D).

Reviewer #3 (Recommendations For The Authors):

Protein homeostasis and lipid homeostasis are both are important for maintaining cellular functions. However, the crosstalk remains largely unknown. The manuscript entitled as "Impairment of lipid homoeostasis causes accumulation of protein aggregates in the lysosome" deals with this interesting topic. An important link between lysosomal protein aggregation and sphingolipids/cholesterol esters metabolism were discovered. The topic belonging to the Cell Biology domain also falls into the aims and scope of eLife. Here are the revisions I recommend:

(1) From lipidomics analysis, a remarkable correlation between levels of sphingomyelin and cholesterol ester and ProteoStat staining was found. Could the authors explain how sphingomyelin and cholesterol ester are quantified? The two lipids are not included as internal standards from the lipidomics experiment.

Sphingomyelin and cholesterol ester internal standards are included in the Avanti 330707 SPLASH® LIPIDOMIX® Mass Spec Standard, which was supplied at 3% v/v to the MeOH/H2O cell lysis buffer. We have amended the Methods section to clarify this.

(2) Could the authors perhaps delete Figure 1B and show it on Figure 2A only? There is no need to show the same figure two times. The threshold of both False Discovery Rate and Median Enrichment needs to be added. From Figure 2A, the Lysosomal hydrolases (GBA, LIPA, GALC) seems located in statistically insignificant region. Based on previous studies, the GBA could have an effect on sphingolipid levels, then how to explain that sphingomyelin was highly correlated with ProteoSate staining?

We have combined the two volcano plots into a single figure (now Figure 1D), and added a line to help visualize the gene effects while considering the combined contribution of FDR and enrichment. Individual lysosomal hydrolases indeed have insignificant effects on ProteoStat and this is discussed in the main text as having relatively constrained impacts on the general lipidome. For example, while GBA and GALC KDs can lead to accumulation of their immediate substrates (glucosylceramide and galactosylceramide, respectively), they do not directly impinge on sphingomyelin.

(3) The authors show the corelation between ProteoState staining and different lipids/lipid classes in Figure 3B and Figure S3A. It is not necessary to show the corelation with individual lipids (such as sphingomyelin(d18:1/24:0) and cholesterol ester(18:2). The corelation with full collection of lipid classes would be more representative, which is only list in Figure 3B and Figure S3A. It is suggested to add the information of how many individual lipids in each chass are used for the correlation analysis. Replace Figure 3A to Figure S3A, and put Figure 3A as supplementary figure are suggested.

We decided to retain the correlation of two individual lipids (a sphingomyelin and a cholesterol ester species) with ProteoStat as examples to illustrate with clarity how we obtained the class-wide comparison. The number of individual lipids included in each class for correlation analysis is now included in Figures 2F and S3A.

(4) The authors state that lipid uptake and metabolism modulate proteostasis. However, only cholesterol and LDL were tested. It would be more precise to state as cholesterol uptake and metabolism modulate proteostasis. In addition, sphingolipids and cholesterol esters accumulate with increased lysosomal protein aggregation. It would be interesting to see the effects of sphingolipids uptake, since sphingolipids are correlated with proteostasis better than cholesterol.

We attempted to add back specific sphingolipids to assess sufficiency. However, we found it challenging to ensure that these lipids were distributed to the correct subcellular locations at physiologically relevant levels. Without this crucial information, it was difficult to draw any conclusions about the sufficiency of the sphingolipids we tested to impair proteostasis.

Alonso A, Goñi FM. 2018. The Physical Properties of Ceramides in Membranes. Annu Rev Biophys 47:633–654. doi:10.1146/annurev-biophys-070317-033309

Ikonen E, Zhou X. 2021. Cholesterol transport between cellular membranes: A balancing act between interconnected lipid fluxes. Dev Cell 56:1430–1436. doi:10.1016/j.devcel.2021.04.025

Niekamp P, Scharte F, Sokoya T, Vittadello L, Kim Y, Deng Y, Südhoff E, Hilderink A, Imlau M, Clarke CJ, Hensel M, Burd CG, Holthuis JCM. 2022. Ca2+-activated sphingomyelin scrambling and turnover mediate ESCRT-independent lysosomal repair. Nat Commun 13:1875. doi:10.1038/s41467-022-29481-4

Version 2

Abstract

Abstract

Protein aggregation increases during aging and is a pathological hallmark of many age-related diseases. Protein homeostasis (proteostasis) depends on a core network of factors directly influencing protein production, folding, trafficking, and degradation. Cellular proteostasis also depends on the overall composition of the proteome and numerous environmental variables. Modulating this cellular proteostasis state can influence the stability of multiple endogenous proteins, yet the factors contributing to this state remain incompletely characterized. Here, we performed genome-wide CRISPRi screens to elucidate the modulators of proteostasis state in mammalian cells, using a fluorescent dye to monitor endogenous protein aggregation. These screens identified known components of the proteostasis network and uncovered a novel link between protein and lipid homeostasis. Increasing lipid uptake and/or disrupting lipid metabolism promotes the accumulation of sphingomyelins and cholesterol esters and drives the formation of detergent-insoluble protein aggregates at the lysosome. Proteome profiling of lysosomes revealed ESCRT accumulation, suggesting disruption of ESCRT disassembly, lysosomal membrane repair, and microautophagy. Lipid dysregulation leads to lysosomal membrane permeabilization but does not otherwise impact fundamental aspects of lysosomal and proteasomal functions. Together, these results demonstrate that lipid dysregulation disrupts ESCRT function and impairs proteostasis.

Introduction

Protein aggregation is a hallmark of aging and age-related pathologies, including multiple neurodegenerative diseases and amyloidoses that affect individual or multiple organs (Chiti and Dobson, 2017; Eisenberg and Jucker, 2012). Dozens of proteins that are normally soluble may form aggregates that are traditionally associated with specific diseases -- e.g. tau tangles and amyloid plaques with Alzheimer’s Disease, and transthyretin deposits with transthyretin amyloidosis (Chiti and Dobson, 2017). Though specific pathogenic proteins are often studied in isolation, a more complex relationship among different aggregation-prone proteins exists, either directly or through interaction with the cellular machinery regulating proteostasis. For instance, experiments in yeast, worms, and mammalian cell culture have demonstrated considerable cross-talk and cross-sensitization between classic aggregation-prone proteins (Gidalevitz et al., 2009; Guo et al., 2013). Further, a diverse set of proteins exhibit reduced solubility and lose their native structure during normal aging (David et al., 2010). Intriguingly, these insoluble age-associated proteins can seed the formation of classical amyloids in vitro, and contribute to functional decline in aging animals (Huang et al., 2019). Together, the evidence suggests that systemic changes in the state of cellular proteostasis may underlie proteinopathy and highlights the importance of expanding the scope of studies beyond specific disease-associated protein species.

The core proteostasis network is well described and involves the synthesis, folding, trafficking and degradation of proteins (Hipp et al., 2019; Kaushik and Cuervo, 2015; Labbadia and Morimoto, 2015). Most of these components closely follow the life cycle of proteins, including 1) pathways that modulate protein synthesis and degradation, such as mTORC, ribosome biogenesis, the ubiquitin proteasome system, and autophagy; 2) molecular chaperones, which aid protein folding; and 3) stress response pathways that regulate the activity of (1) and (2) in response to accumulation of unfolded or misfolded proteins, such as ER-associated degradation and the unfolded protein response. Nevertheless, the components and network topology vary depending on cell type and cell state. Furthermore, mechanisms that underlie shifts in cellular proteostasis state, especially ones that do not directly involve core nodes of the proteostasis network, remain largely unknown.

Unbiased, whole-genome screening is a powerful tool for discovering cellular components and pathways that are involved in modifying a specific phenotype. Previous studies have applied genetic screens in different cell models to uncover, for specific proteins, modifiers of their aggregation, toxicity, or propensity for aggregation to propagate from seeds. These systems are often driven by overexpressing aggregation-prone proteins and/or seeding with extracellular pre-formed fibers (Kramer et al., 2018; Chen et al., 2019; Willingham et al., 2003). Thus, while much progress has been made in understanding the homeostasis of several discrete proteins, the mechanisms influencing the broader cellular proteostasis network remain largely unknown. In particular, we are interested in cellular mechanisms that affect the aggregation state of endogenously expressed proteins, as these may be the most germane to age-associated loss of proteostasis.

To uncover novel components and pathways that regulate endogenous protein homeostasis, we applied CRISPR interference (CRISPRi) technology to perform unbiased, genome-wide screens on K562 cells using an aggresome-staining dye. ProteoStat is known to label both experimentally induced aggresomes (Shen et al., 2011) and aggregates that accumulate with age in the CNS of mammals (Leeman et al., 2018; Gefen et al., 2015; Vonk et al., 2020). To our knowledge, ours are the first CRISPRi genome-wide screens performed on unmodified mammalian cells to interrogate endogenous protein aggregation. These screens revealed a link between lysosomal lipid homeostasis and proteostasis. Increased levels of sphingomyelins and cholesterol esters, either as a result of increased lipid uptake or decreased lysosomal degradative capacity, lead to increased protein aggregation predominantly colocalized with the lysosomes. Proteomic analysis of isolated lysosomes revealed accumulation of detergent-insoluble ESCRT components, indicating potential impairment of lysosomal membrane repair and microautophagy. Lipid dysregulation induces or potentiates lysosomal membrane permeabilization (LMP), the rescue of which partially restores proteostasis. Other aspects of lysosomal and proteasomal functions essential for proteostasis (lysosomal pH, lysosomal protease activity, and proteasome activity) were unaffected by lipid dysregulation. Together, this work provides strong evidence that supports an emergent role of lipid dysregulation in driving protein aggregation through disruption of the ESCRT pathway.

Results

Measurement of endogenous proteostasis state by flow cytometry

ProteoStat is used to stain endogenous protein aggregates in fixed cells. Like Thioflavin T, ProteoStat fluorescence increases when rotationally confined by binding to cross-beta sheets in amyloidal structures. Recent studies have used the dye to visualize aggregated proteins in mouse NSCs and C. elegans (Leeman et al., 2018; Shen et al., 2011; Vonk et al., 2020; Bohnert and Kenyon, 2017). To verify that ProteoStat can report changes in proteostasis in human cell culture, we inhibited proteasomes in K562 cells and stained them for microscopy or flow cytometry analysis. As expected, proteasome inhibition by MG132 increased ProteoStat intensity, with numerous bright puncta around the nucleus (Figure 1–figure supplement 1A). ProteoStat-high cells obtained by fluorescence-activated cell sorting (FACS) from cells treated with a lower dose of MG132 -- which generated mixed phenotypes -- likewise showed numerous bright puncta, whereas ProteoStat-low cells displayed diffuse staining of lower intensity, indicating that this tool can be used to report and isolate cells in different proteostasis states (Figure 1A).

CRISPRi screens identified major cellular proteostasis components and implicated lipid uptake and metabolism pathways

A) ProteoStat validation: MG132-treated cells were stained, sorted into top and bottom quartiles (ProteoStat-high and ProteoStat-low, respectively), and imaged. B) CRISPRi screen schematic: Untreated K562 cells were infected with sgRNA library, selected, stained, and sorted using gates that corrected for correlation between staining intensity and cell size. C) A subset of GO Terms (Biological Process and Cellular Component) identified with GSEA on screen results. Positive Enrichment scores indicate an increased likelihood of genes in the GO Term to increase ProteoStat when knocked down. D) Volcano plot highlighting major proteostasis pathways and select genes involved in lipid uptake and metabolism. Dotted line denotes where the product of |log2(fold-change)| and -log10(False discovery rate) is constant and equals to 1*-log10(0.05).

A) K562 cells treated with MG132 at 1 uM for 20 hours were stained with ProteoStat and imaged. Images with untreated and treated cells are shown here with the same grayscale. Inset: enlarged image of a single cell showing perinuclear ProteoStat punctates in MG132-treated cells. B) Results of a reduced-scale batch retest to validate the effect of select sgRNAs from the genome-wide screen. Sequencing counts of each sgRNA in the FACS-sorted ProteoStat-high vs ProteoStat-low populations are shown. sgRNAs selected include ones that were enriched in the ProteoStat-high (red) and ProteoStat-low (blue) populations, or showed no enrichment (green) in the genome-wide screen, as well as non-targeting sgRNAs (gray) for normalization. C) Effects of pharmacological inhibition of proteasome (MG132, 1 uM), lysosomal acidification (Bafilomycin A1, 6.25 uM), and chaperones (VER-155008, 50 uM; Ganetespib, 100 nM; 17-DMAG, 1 uM; n=1) on ProteoStat staining in K562 cells after the specified treatment durations.

Genome-wide screen for modifiers of ProteoStat recovered major known components of the proteostasis machinery

We performed FACS-based, pooled screens in K562 CRISPRi cells, which stably expressed KRAB-dCas9. Briefly, cells carrying individual sgRNAs were fixed, stained, and sorted by ProteoStat intensity into top and bottom quartiles (ProteoStat-high and ProteoStat-low, respectively; Figure 1B). Illumina sequencing was used to quantify the frequency of each sgRNA in each of these two populations. We performed this experiment twice and evaluated the effect of each gene’s knockdown (KD) on proteostasis using the MAGeCK package (Li et al., 2014).

KDs of 239 genes increased ProteoStat, whereas KDs of 607 genes decreased it (out of 20,704 protein-coding genes targeted by the sgRNA library; FDR < 0.05). We validated the results of a subset of genes by repeating the screen at a reduced scale (batch retest; Figure 1–figure supplement 1B). Gene set enrichment analysis (GSEA) captured ontologies in which genes showed consistent effects in the same direction, even though individual genes may fall below an artificial threshold. Many of these genes and pathways are known proteostasis modifiers (Figure 1C-D). These include ribosomal proteins (both small 40S and large 60S subunits), ribosome biogenesis and processing, proteasome subunits, lysosomal proton pump subunits (vATPases), and certain chaperones. KD of many ribosomal proteins and ribosome assembly factors decreased ProteoStat, consistent with reduced input of proteins into the proteostasis network (Stein and Frydman, 2019). Knocking down subunits of the proteasome complex increased ProteoStat, consistent with MG132 treatment in our validation experiments. Disrupting vATPases similarly increased ProteoStat, likely via deacidification and impairment of lysosomes, an important center where multiple catabolic processes converge. These effects on ProteoStat following inhibition of proteasome and lysosomal acidification were replicated using pharmacological inhibitors (Figure 1–figure supplement 1C). In contrast, KD of many chaperones, including HSPA4, HSP90AB1, DNAJB6, and several subunits of TRiC chaperonins, paradoxically reduced ProteoStat. Perturbing chaperones can trigger the newly described safeguard against protein aggregation mechanism that targets certain newly synthesized aggregation-prone proteins to the lysosome, notably before large visible aggregates are formed (Jung et al., 2020). Alternatively, initial HSP70 or HSP90 inhibition can upregulate heat shock response as a negative feedback (Kijima et al., 2018; Åkerfelt et al., 2010; Ali et al., 1998). Indeed, pharmacological inhibition of chaperones increased ProteoStat after 30 minutes, but decreased after 24 hours (Figure 1– figure supplement 1C).

Surprisingly, KD of most of the major components directly involved in macroautophagy did not affect ProteoStat in either direction. These include the core autophagosome formation components (e.g. ATG5, ATG10, ATG12, and the ATG8 protein family) and the autophagy adaptors (NBR1, TAX1BP1, OPTN, SQSTM1 etc). Though we performed our screens under nutrient-replete conditions, perturbations of these macroautophagy genes in K562 under similar growth conditions have previously been demonstrated to influence autophagic flux (Shoemaker et al., 2019). These results suggest that basal macroautophagy plays a negligible role in clearing endogenous amyloid-like structures in this system.

Lipid uptake and metabolism modulate proteostasis

The strong association of lysosomal vATPases with Proteostat phenotype is likely due to the need to maintain a lysosomal pH of ∼4.5-5.0 for optimal activity of the hydrolases that break down proteins, lipids, and nucleic acids. Failure to maintain this acidic pH can cause lysosomal storage of undegraded macromolecules. Accordingly, in addition to the vATPases, several genes that broadly regulate lysosomal function impacted ProteoStat. These include genes that modulate the delivery or activity of multiple lysosomal enzymes (e.g. GNPTAB, IGF2R, LYSET, and PSAP) or lipid uptake and metabolism (e.g. LDLR, MYLIP, and SCAP) (Figure 1D). However, no perturbations of more discrete activities (e.g. individual cathepsins, lipid hydrolases -- e.g. GBA, LIPA, and GALC -- or solute carriers) did so, suggesting that proteostasis is generally resilient and becomes impaired only in response to pleiotropic perturbations.

KD of low-density lipoprotein receptor (LDLR) reduced ProteoStat. Correspondingly, KD of the E3 ubiquitin-protein ligase MYLIP, which targets LDLR for sterol-regulated degradation, led to one of the strongest signal increases. MYLIP KD increased ProteoStat via upregulating LDLR levels rather than through other MYLIP targets, as the effect of LDLR KD was epistatic to MYLIP in cells with double KD (Figure 2A). Cultured cells take up LDL - lipoproteins rich in cholesterol esters - through LDLR and the endo-lysosomal pathway. Once inside the acidic environment of lysosomes, cholesterol ester is hydrolyzed into free fatty acid and cholesterol, the latter of which is trafficked to the ER. Excess cholesterol exerts negative feedback on LDLR levels via two mechanisms: 1) oxysterols (oxidized derivatives of cholesterol) activate the liver X receptor (LXR), in turn increasing MYLIP expression and LDLR degradation (Zelcer et al., 2009); 2) high ER cholesterol sequesters the SCAP-SREBPs complex, suppressing its activity and in turn reducing LDLR expression as well as de novo cholesterol synthesis (Goldstein et al., 2006). Two positive regulators of the SREBP pathways (SCAP and MBTPS2) also reduced ProteoStat when knocked down (Figure 1D). In contrast, genes along the mevalonate -- de novo cholesterol synthesis -- pathway did not consistently score in our screen, indicating the difference between cholesterol that is taken up vs synthesized de novo.

Accumulation of sphingolipid and cholesterol ester can impair proteostasis

A & D) Comparing effects on ProteoStat of single vs double KDs (n ≥ 2) B) Effects of gene KDs on ProteoStat in cells maintained in media with regular or delipidated FBS (DL media). (n = 3) C) Effects of supplementing free cholesterol or human LDL in restoring ProteoStat, which was lowered in cells grown in DL media (n = 3). E) ProteoStat vs abundance of indicated lipid species in cells with KD of a core set of lipid-related target genes, or including other ProteoStat perturbing genes (insets). Both ProteoStat and lipid abundance are averages of biological triplicates and normalized to cells carrying non-targeting guides. Least squares linear regression models (dotted red lines) were generated and the Coefficient of determination, R2, values are shown. F) Histograms of R2 values for all lipid species within each indicated lipid class as in E. p-values were calculated using K-S test comparing R2 values within each lipid class versus that of all lipids. The number of individual lipid species analyzed is indicated for each lipid class. Phenotypes in A-D were normalized first to an internal staining control, and then to cells carrying non-targeting guides. (✻: p<0.05; ✻✻: p<0.01; t-test between pairs of samples as indicated with horizontal lines)

A & B) Filipin III (A) and Proteostat (B) staining on K562 cells that were untreated or MBCD-treated to remove cholesterol. Same grayscale was used for each pair of images in the main panels, where an additional Filipin III image of the MBCD-treated cells was shown at enhanced-grayscale to expose the minimal filipin III staining on these cells. C) Co-staining K562 cells (untreated or MG132-treated) against LAMP1 and with ProteoStat and Filipin III. D) Biphasic GUVs with cholesterol-poor and cholesterol-rich phases were generated, which included TopFluor Cholesterol to label the cholesterol-rich phases. These GUVs were stained with ProteoStat at two concentrations -- low (1:10000 diluted, as used in all other experiments, including FACS-based screen) and high (1:500 diluted). Unstained samples were included to assess background fluorescence in the ProteoStat channel.

A) Histograms of R2 values obtained for all lipid species within each indicated lipid class --- ceramides (Cers), lactosylceramides (LacCers), or hexosylceramides (HexCers) --- using linear regression models generated from the ProteoStat phenotypes vs lipid abundances, as in Figure 2E. p-values were calculated using K-S test comparing R2 values within each lipid class versus that of all lipids. The number of individual lipid species analyzed is indicated for each lipid class. B) Excerpt of hydrolytic pathway of a few sphingolipid species, their catalytic enzymes, and associated activating Saps. C) Schematic of the series of constructs expressing different variations of PSAP (full-length, truncated, carrying mutations deleterious to individual Sap peptides, or individual functional Sap peptides alone). Colored boxes depict individual Sap peptides. Deleterious mutations are depicted by X’s on the corresponding Sap peptide. D) ProteoStat phenotypes in cells overexpressing various PSAP variants in E, in conjunction with either PSAP KD or non-targeting control (n=3, #: p>0.05; t-test).

Chronic depletion of exogenous LDL using media that contains delipidated serum reduced ProteoStat to a level similar to LDLR KD and was sufficient to abrogate the effect of MYLIP KD (Figure 2B), further demonstrating that the effect of these genetic perturbations was mediated through lipid uptake and not direct degradative action of MYLIP on putative aggregation-prone substrates. Supplementing either exogenous cholesterol or human LDL reversed the effect of LDL depletion (Figure 2C). We also examined the possibility that ProteoStat may directly stain cholesterol by acutely extracting cholesterol using methyl-β-cyclodextrin (MBCD). MBCD removed most free cholesterol as determined by Filipin III staining, but did not affect ProteoStat (Figure 2–figure supplement 1A-B). In neither untreated nor MG132-treated cells did ProteoStat and Filipin III staining overlap, indicating that ProteoStat does not stain free cholesterol or any major components of the plasma membrane (Figure 2–figure supplement 1C). Furthermore, in an in vitro biphasic lipid bilayer system (giant unilamellar vesicles, GUVs), ProteoStat stained the lipid bilayer poorly and was not brighter at the cholesterol-rich phase (Figure 2–figure supplement 1D). Together, the evidence suggests that increased flux of LDL uptake and/or the levels of lipid metabolites generated during the process increases protein aggregation.

KD of genes involved in lysosomal sphingolipid metabolism and lysosomal protein trafficking (GNPTAB, IGF2R, LYSET and PSAP) increased ProteoStat in our screens (Figure 1D). GNPTAB encodes two subunits of a Golgi-resident enzyme, GlcNAc-1-phosphotransferase, which catalyzes the first step to form mannose 6-phosphate (M6P) markers on lysosomal enzymes and is stabilized by the recently characterized LYSET (also named TMEM251) (Richards et al., 2022). IGF2R (also known as the cation-independent M6P receptor) binds to these lysosomal enzymes and facilitates their transport from golgi to lysosome. PSAP, or prosaposin, is the precursor to saposin peptides, which are required to activate lysosomal sphingolipid catabolizing enzymes. The effect of PSAP KD was dramatically exacerbated by MYLIP KD and alleviated by LDLR KD (Figure 2D), indicating that lipid uptake can impinge on proteostasis through its interaction with lipid catabolism in lysosomes. The individual lysosomal hydrolases activated by saposins, such as GBA, GALC, and LIPA, did not rise above the detection threshold in our screen, again suggesting that the proteostasis network, as measured by ProteoStat, is generally resilient to perturbations of discrete functional nodes or changes in individual lipid species.

Sphingolipids and cholesterol esters mediate the effect of lipid perturbations on proteostasis

Having demonstrated that lipid metabolism could influence proteostasis, we wondered if specific lipid species might be involved. To address this question, we performed lipidomic analysis on cells in which lipid metabolism was perturbed, either genetically by CRISPRi, or environmentally by lipid depletion. These perturbations were selected from the screen results to represent a range of effects on ProteoStat staining. When considering perturbations that directly affect lipid uptake and metabolism, we found remarkable correlation between ProteoStat staining and individual levels of most lipids in the sphingomyelin (SM) and cholesterol ester (CE) classes (R2of up to 0.968 and 0.843, for SM and CE, respectively, Figure 2E). Indeed, lipids belonging to these two classes were better correlated with the ProteoStat phenotype when compared with the full collection of lipid compounds assayed (Figure 2F, see Figure 2–figure supplement 2A for the lipid classes ceramides (Cers), lactosylceramides (LacCers), or hexosylceramides (HexCers) as counterexamples). In contrast, KD of targets that do not impinge directly on lipid metabolism (e.g. AMBRA1 or CHMP2A) or else expected to have strong pleiotropic effects beyond lipid metabolism (e.g. ATP6V1E1) did not induce strong correlations between SMs/CEs and ProteoStat.(insets in Figure 2E). These results further highlight the multimodality of proteostasis regulation, where lipid dysregulation represents one, among other, major components.

To determine if any lipid species alone contributed critically to protein aggregation induced by lipid dysregulation, we leveraged the modularity of the highly conserved PSAP. Upon delivery to the lysosome, the proprotein is cleaved proteolytically into four active peptides, saposins (Saps) A, B, C, and D. Saps are involved in the hydrolysis of glycosphingolipids (GSLs) as enzymatic activators, facilitating the interaction between soluble lysosomal hydrolases and GSLs embedded in the membrane (Kolter and Sandhoff, 2005; Xu et al., 2010). Distinct residues within the Sap-hydrolase interface confer hydrolase specificity on each saposin (e.g. SapA binds to Galactocerebrosidase, GALC, and stimulates galactosylceramide hydrolysis, Figure 2–figure supplement 2B) (Hill et al., 2018). Deficiency in full-length PSAP or individual Saps (by truncation, inactivating mutations, or reduced expression) lead to various lysosomal storage disorders, with neuropathology and often dysfunction of other organs (Schulze and Sandhoff, 2011). Depending on which Sap is deficient, distinct GSL species are found to accumulate in the storage cells. We therefore performed rescue experiments using different PSAP variants to perturb turnover of individual GSLs more specifically. We expressed full length PSAP or variants with either a single Sap or all-but-one Sap (Figure 2–figure supplement 2C). Full-length PSAP largely rescued the increase in ProteoStat induced by PSAP KD (p=0.07 vs non-targeting, no-rescue control), while the variant with all four Saps inactivated (PSAP ΔABCD) failed to rescue (p=0.74 vs PSAP KD, no-rescue control; Figure 2–figure supplement 2D). Furthermore, addition of any single Sap did not fully rescue the KD phenotype. Accordingly, inactivating any single Sap also did not fully abrogate PSAP’s ability to rescue. Together, the evidence suggests that proteostasis impairment from sphingolipid accumulation is pleiotropic and is likely attributable to the action of multiple but not a single lipid. This is consistent with the observation that, in our CRISPRi screens, none of the individual GSL hydrolases significantly increased ProteoStat level.

Lipid dysregulation leads to accumulation of insoluble lysosomal protein aggregates

Lipid-related perturbations that impair proteostasis suggested a central role for lysosomes, prompting us to further characterize ProteoStat staining, especially its subcellular localization. Aggresomes are reportedly composed of misfolded and aggregated proteins that are poly-ubiquitinated and transported along microtubules to coalesce into perinuclear assemblies. These aggresomes can be recognized by autophagy receptors to undergo canonical macroautophagy and lysosomal degradation (Kocaturk and Gozuacik, 2018). Alternatively, cytosolic or organelle cargo may be recruited and directly delivered into endolysosomal compartments in a process termed microautophagy that is dependent on the ESCRT (endosomal sorting complex required for transport) pathway (Kuchitsu et al., 2023; Oku et al., 2017; Schäfer et al., 2020). ProteoStat was initially reported to colocalize with mono-/poly-ubiquitinated proteins, p62, and LC3, and has been used as a general reporter for protein aggregation (Bohnert and Kenyon, 2017; Leeman et al., 2018; Shen et al., 2011; Vonk et al., 2020). Additionally, lysosome-accumulating ProteoStat-positive aggregates were reported in quiescent neural stem cells but not their activated counterparts (Leeman et al., 2018), nor in other rapidly dividing cell types.

We initially characterized aggresomes induced by pharmacological proteasome inhibition. Treatment with MG132 for four hours induced ProteoStat puncta that colocalized predominantly with the lysosomes, in addition to a diffuse but weaker signal in the cytosol (Figure 3–figure supplement 1A-B). A similar distribution was observed after 20 hours of treatment. In contrast, mono-/poly-ubiquitinated proteins increased only modestly at 4 hours; at 20 hours these signals increased globally and did not concentrate in the lysosomes. Punctate structures of mono-/poly-ubiquitinated proteins became more apparent at a higher dose of MG132. These puncta were non-lysosomal and ProteoStat-positive, though their ProteoStat intensities were lower than the lysosomal ProteoStat puncta (Figure 3–figure supplement 1A, white arrowheads). In comparison, proteasome inhibition by Bortezomib led to proportionally stronger cytoplasmic ProteoStat and mono-/poly-ubiquitination and less pronounced lysosomal puncta. Finally, when lysosomes were deacidified with Bafilomycin A1, we observed strong overlapping signals of ProteoStat and mono/poly-ubiquitination that were partially colocalized with the lysosomes, consistent with impaired ability to remove and degrade ubiquitin chains or ubiquitinated proteins. Patterns of Thioflavin T staining were similar to ProteoStat (Figure 3–figure supplement 1C-D). These results with diverse inhibitors demonstrate subtle pharmacological biases – with MG132 favoring lysosomal puncta and Bortezomib causing both classical aggresomes and lysosomal phenotypes – and the ability for ProteoStat to capture these variations. These observations of lysosomal ProteoStat-high ubiquitination-low puncta, together with the aforementioned lack of evidence of macroautophagy involvement, are consistent with the microautophagy pathway where cargos are processed by deubiquitinating enzymes (STAMBP and UBPY) associated with ESCRT during transport into lysosomes (Hanson et al., 2009). Under mild proteostatic stress, ubiquitinated proteins accumulate and are transported into lysosomes by microautophagy, while being deubiquitinated in the process. When proteostatic stress prolongs or intensifies, deubiquitinating capacity may be overwhelmed, resulting in increased accumulation of ubiquitinated proteins both inside and outside lysosomes. Indeed, UBPY KD increased lysosomal signals of both ProteoStat and mono-/poly-ubiquitinated proteins, consistent with a role for microautophagy in delivery of misfolded proteins to the lysosome (Figure 3–figure supplement 1E-F).

Cells with lipid impairment (MYLIP or PSAP KD) exhibited lysosomal ProteoStat puncta that were non-ubiquitinated, in addition to occasional cytoplasmic double-positive puncta (Figure 3A-B, white arrowheads). This suggests a moderate level of proteostatic stress that does not overwhelm the capacity to deubiquitinate. Accordingly, cells with MYLIP and PSAP KD did not exhibit growth defects (Figure 3– figure supplement 2A). To verify the presence of protein aggregates in lysosomes under the above conditions, we isolated lysosomes using Lyso-IP (Lysosome Immunoprecipitation) (Abu-Remaileh et al., 2017). HA epitopes are targeted to the lysosomal surface by TMEM192, which enable lysosomal isolation, followed by differential extraction into RIPA (detergent-soluble) and Urea/SDS (detergent-insoluble) fractions. This orthogonal approach confirmed that lipid-perturbation (MYLIP and PSAP KDs) resulted in accumulation of detergent-insoluble protein aggregates in the lysosome (Figure 3C). Inhibiting lysosomal acidification (ATP6V1E1 KD) or proteasome activity (by MG132) led to larger amounts of lysosomal aggregates (Figure 3C & Figure 3–figure supplement 2B).

Lipid dysregulation leads to accumulation of insoluble lysosomal protein aggregates

A) K562 cells with various KDs were co-stained with ProteoStat and antibodies against ubiquitinated proteins or lysosomes and imaged with confocal (maximum projection of multiple-z-slices). B) Quantification (n=3) of ProteoStat and mono/poly-ubiquitinated proteins (representative images shown in A) segregated according to co-localization with lysosome (LAMP1 staining). C) Representative (from triplicate experiments) polyacrylamide gel with Silver Stain showing Lysosomal Insoluble fractions from cells with various KDs after Lyso-IP and differential extraction. (✻✻: p<0.01; n.s.: not significant; t-test vs NT control or between pairs of samples as indicated with horizontal lines)

A & C) K562 cells with various treatments were co-stained with ProteoStat and antibodies against ubiquitinated proteins or lysosomes (A), or Thioflavin T and antibodies against lysosomes (C), and imaged with confocal microscopy. White arrowheads highlight cytoplasmic Ub-high/ProteoStat-medium puncta. B & D) Quantification (n=3) of ProteoStat, mono/poly-ubiquitinated proteins (B), or Thioflavin T (D) signals, segregated according to co-localization with lysosome (representative images of select conditions shown in A & C, respectively). E) K562 cells with control or UBPY KD were co-stained with ProteoStat and antibodies against ubiquitinated proteins or lysosomes. F) Quantification (n=3) of ProteoStat and mono/poly-ubiquitinated proteins (representative images shown in E) segregated according to co-localization with lysosome (LAMP1 staining). Maximum projection of multiple z-slices shown for all microscopy images.

A) Effect on growth rates of various gene KDs. B) Polyacrylamide gel with Silver Stain showing Lysosomal Insoluble fractions from cells treated with various concentrations of MG132 for 20 hours, followed by Lyso-IP and differential extraction.

Lipid dysregulation leads to the accumulation of insoluble ESCRT components at the lysosome

We performed in depth quantitative proteomics coupled with isobaric tandem mass tags (TMT) on different Lyso-IP fractions to determine the identity of lysosomal detergent-insoluble proteins. 67 of 70 core lysosomal resident (Singh et al., 2020) proteins were captured in the Lysosomal Total fraction (before differential extraction), validating the Lyso-IP-proteomics method. Within the Lysosomal Insoluble fraction, 35 proteins were over-represented (q-value <0.05) in MYLIP KD cells versus control. Remarkably, 9 of the top 12 insoluble proteins induced by MYLIP KD were components involved in the ESCRT pathway (Figure 4A), including multiple ESCRT-III subunits (e.g. CHMP2A, CHMP4A & B), accessory proteins (ALIX and PTPN23), and ESCRT-III disassembling enzymes (VPS4A & VTA1). The ESCRT machinery is involved in multiple cellular processes that involve membrane scission with a negative curvature (membrane bending out and away from the cytoplasm), including cytokinesis, ciliogenesis, multivesicular body formation, microautophagy, and lysosomal membrane repair. Accordingly, GSEA of insoluble lysosomal proteins enriched in MYLIP KD compared to control highlighted the Multivesicular Body gene set (GO:0005771) (Figure 4B). This enrichment of ESCRT components was not simply a consequence of increased abundance in the Lysosomal Total fraction, as multiple proteins (most notably vATPases) were also more abundant in the total, but not insoluble, fraction of MYLIP KD cells (Figure 4–figure supplement 1A). ESCRT-III subunits have previously been shown to form detergent-insoluble hetero-polymers on membranes (Hanson et al., 2008; McCullough et al., 2015). These ESCRT-III polymers can be trapped on membranes by depleting VPS4A/B (Cashikar et al., 2014), which has the dual function of progressing ESCRT-III polymerization through subunit exchange and ultimately disassembling the complex upon completion of membrane scission (Pfitzner et al., 2020). Here, our results provide novel evidence that excessive SM and CE levels may disrupt ESCRT function by impeding ESCRT-III progression and/or resolution.

Lysosomal protein aggregates under lipid accumulation are enriched in ESCRT components

A) Volcano plot of Lyso-IP proteomics experiment highlighting ESCRT components and associated proteins that became more insoluble in the lysosome in MYLIP KD vs control cells. B) GO Terms (Cellular Component) identified with GSEA on Lyso-IP Proteomics results. Positive Enrichment scores indicate an increased likelihood of genes in the GO Term to be enriched in Lysosomal Insoluble fraction in MYLIP KD vs control cells. Dotted line denotes where the product of |log2(fold-change)| and -log10(False discovery rate) is constant and equals to 1*-log10(0.05).

A) Relative enrichment of proteins in MYLIP KD over NT control, in Lysosomal Total Fraction vs Lysosomal Insoluble Fraction. Highlighted are proteins with statistically significant enrichment in various categories.

Lysosomal membrane permeabilization (LMP) contributes to proteostasis impairment

Can ESCRT disruption mediate the effect of lipid dysregulation on proteostasis impairment? As ESCRT contributes to lysosomal membrane repair and ESCRT deficiency is sufficient to drive LMP, we examined the effect of lipid-perturbing KDs on lysosomal membrane integrity using a TagBFP-fused Galectin-3 (GAL3) reporter. GAL3 is normally diffused in the cytoplasm but becomes punctate upon binding to glycans on the lumenal side of the lysosomal membrane that are exposed upon damage. We imaged the reporter under a combination of candidate gene KDs and/or acute LMP induced by LLOMe (L-leucyl-L-leucine, a lysosomotropic membranolytic agent) (Uchimoto et al., 1999). MYLIP KD and, to a much weaker degree, PSAP KD, both increased LMP, as did KDs of ESCRT-III subunits (CHMP2A and CHMP6) and ATP6V1E1 (Figure 5A-B). Chemically inducing LMP with LLOMe was sufficient to cause a moderate lysosomal accumulation of protein aggregates (Figure 5C). However, the magnitude of LMP was not directly related to ProteoStat intensity across conditions (Figure 5C-E), suggesting that while LMP is sufficient to promote lysosomal protein aggregation, other factors are likely to contribute.

Lysosomal membrane permeabilization contributes to proteostasis impairment

A) Representative fluorescence microscopy images of K562 cells expressing TagBFP-GAL3, with NT control, sgRNA targeting MYLIP, or treated with LLOMe at 120 uM for 4 hours. B) Quantification (n=3) of TagBFP-Gal3 puncta in cells with various KDs. C) Quantification (n=3) of lysosomal ProteoStat signal under various treatments. D) Quantification (n=3) of TagBFP-Gal3 puncta in cells with a combination of control vs MYLIP KD, ±LLOMe treatment (120 uM for 4 hours), and ±VPS4A overexpression. E) Quantification (n=3) of lysosomal ProteoStat signal in cells with a combination of control vs MYLIP KD and ±VPS4A overexpression. (✻: p<0.05; ✻✻: p<0.01; n.s.: not significant; t-test vs NT / Untreated control or between pairs of samples as indicated with horizontal lines)

We further tested if LMP was necessary for lysosomal protein aggregation by promoting lysosomal membrane repair through VPS4A overexpression. Chemically induced LMP was almost completely rescued by ectopic VPS4A expression (Figure 5D), presumably due to accelerated progression and resolution of ESCRT-III polymers. This rescue of chemically induced LMP was largely negated in the background of MYLIP KD (Figure 5D), suggesting that lipid dysregulation impairs ESCRT pathway in an orthogonal manner or upstream to ESCRT-III resolution. Accordingly, VPS4A overexpression only led to a modest reduction in LMP and lysosomal protein aggregation that were induced by MYLIP KD (Figure 5D-E). Together, these data show that LMP contributes, albeit only partially, to lipid-induced lysosomal protein aggregation and that lipid accumulation may impair VPS4A-mediated resolution of ESCRT assemblies

Proteostasis impairments due to lipid dysregulation are not correlated with changes in lysosomal and proteasomal functions

We sought to determine whether lipid dysregulation impacted proteasomal or lysosomal function. SM and CE metabolisms were previously linked with several diseases with proteinopathy, including Alzheimer’s disease (AD) and the lysosomal storage disease Niemann Pick’s type C (NPC) (Kirkegaard et al., 2010; Schulze and Sandhoff, 2011; Van Der Kant et al., 2019; Yamazaki et al., 2001). Furthermore, multiple lines of recent evidence suggest that SM and/or CE dysregulation may induce lipid peroxidation (Choi et al., 2017; Tian et al., 2021) and impinge on the activity of the proteasome and lysosomal proteases (Gabandé-Rodríguez et al., 2014; Van Der Kant et al., 2019). We therefore investigated different aspects of lysosomal health, including lysosomal number, size, protease activity, pH, and membrane integrity, as well as lipid peroxidation and proteasome activity, under several lipid perturbing conditions.

To monitor lysosomal protease activity, we incubated cells with BSA conjugated to self-quenched red BODIPY TR-X (DQ-Red BSA). Uptake and delivery of these molecules to the lysosome lead to BSA hydrolysis and unquenching of BODIPY fluorescence. As expected, impeding lysosomal acidification (by ATP6V1E1 KD) or lysosomal membrane integrity (by CHMP2A KD) led to reduced BODIPY unquenching (Figure 6A). However, despite inducing LMP to a similar degree as KDs of ATP6V1E1, CHMP2A, and CHMP6 (Figure 5B), lipid-modifying perturbations affected lysosomal protease function only modestly and incoherently with respect to the ProteoStat phenotype, with MYLIP KD reducing BODIPY unquenching and PSAP KD trending in the opposite direction (Figure 6A).

Lysosomal membrane permeabilization (LMP) contributes to proteostasis impairment

A-D) Flow cytometry quantification of phenotypes in K562 cells with various KDs (n=3 for all). A: TR-X fluorescence (after protease-induced unquenching) as an indicator of lysosomal protease function; B: Ratio of lysosome-targeted sfGFP fluorescence (pH sensitive) to mCherry fluorescence (pH insensitive) as an indicator of lysosomal pH; C: LysoTracker-Red fluorescence as an indicator of lysosomal content and pH; D: LiperFluo staining as an indicator of lipid peroxidation level; E-F) Efficacy of the 5 antioxidant cocktail (5AO) in rescuing the effects of gene KDs on E) LiperFluo or F) ProteoStat phenotypes. (✻: p<0.05; ✻✻: p<0.01; t-test vs NT control or between pairs of samples as indicated with horizontal lines; n=3) G) Flow cytometry-based assay with the Proteasome Activity Probe was validated with proteasome inhibitors (n=1). H) Effect of gene KDs on proteasome activity (n=1).

A) K562 cells stably expressing the pHLys sensor pHLARE were incubated with a pH buffer series in the presence of ionophores (nigericin and monensin) and analyzed with flow cytometry (n=3). B) Effect of target gene KDs on LysoTracker vs LiperFluo phenotypes. R2 value was obtained with a linear regression model (dotted blue line). C) LysoTracker-Red staining in K562 CRISPRi cells carrying sgRNAs targeting various genes, which were either untreated or treated with the 5 antioxidant cocktail (5AO) for the duration of the gene KD. D) SDS-PAGE in-gel fluorescence of lysates from cells treated with MG132 or Bortezomib for 1, 2, 4, or 20 hours before labeling with the Proteasome activity probe for 1 hour followed by lysis. Expected migration distances of different proteasome subunits are marked. E) Effect of KDs of select genes in the mevalonate pathway on proteasome activity (n=1). F) Transmission Electron Microscopy (TEM) images of K562 CRISPRi cells harvested 6 days after transduction of either non-targeting or PSAP-targeting sgRNAs.

Changes in lysosomal pH (pHLys) can be detected with high sensitivity using a ratiometric reporter (pHLARE), which consists of a pH-sensitive fluorescent protein (super-folder GFP, sfGFP) and a pH-insensitive one (mCherry) flanking rat LAMP1 (Webb et al., 2021). Here, we stably expressed pHLARE, and validated the sensor for use with flow cytometry using a buffer series with pH ranging from 4 to 7.5 in the presence of ionophores (Figure 6–figure supplement 1A). The sensor also correctly reported lysosomal deacidification as expected with ATP6V1E1 KD, and demonstrated that LMP induced by either CHMP2A or CHMP6 KD was associated with modest increase in pHLys (Figure 6B). Nevertheless, lipid-modifying KDs -- MYLIP, PSAP, and LDLR -- did not significantly shift pHLys, suggesting their effects on proteostasis were not mediated via lysosomal deacidification.

We investigated the level of lipid peroxidation and lysosomal content (size and/or number) under lipid-modifying perturbations. Using a fluorescent indicator, LiperFluo, we found that perturbations that increased or decreased SM and CE levels (MYLIP and PSAP KDs, vs LDLR KD) increased or decreased lipid peroxidation, respectively (Figure 6D). Interestingly, these lipid peroxidation phenotypes mirrored changes in total LysoTracker signal (Figure 6C-D, Figure 6–figure supplement 1B). As we did not observe these perturbations substantially altering pHLys (Figure 6B), the LysoTracker result suggested changes in lysosomal content, potentially to compensate for lysosomal impairment that may or may not be a consequence of lipid peroxidation (Kirkegaard et al., 2010). We tested if lipid peroxidation is necessary for the increased protein aggregation and lysosomal content using media with antioxidants. A cocktail of five antioxidants (5AO) is commonly included in the media for culturing neurons – a cell type that is particularly sensitive to oxidative stress. While 5AO was effective in reducing lipid peroxidation, it did not rescue the Proteostat or Lysotracker phenotypes (Figure 6E-F; Figure 6–figure supplement 1C). We therefore conclude that protein aggregation and lysosomal content increase are likely upstream or independent of lipid peroxidation.

Finally, we investigated if the other major arm of protein catabolism, the proteasome, is regulated by changes in lipid metabolism. A recent report suggested that high CE levels in cells may hinder phospho-Tau degradation due to proteasome inhibition (Van Der Kant et al., 2019). We labeled active proteasomes using a fluorescent probe, Me4BodipyFL-Ahx3Leu3VS (Me4BoVS) (Berkers et al., 2012, 2007), which covalently binds to catalytically active proteasome β-subunits in live cells. MG132 and Bortezomib treatments reduced the fluorescent signal from Me4BoVS in both live cell flow cytometry and in-gel analysis of cell lysates, validating its use as a proteasome activity probe (Figure 6G, Figure 6–figure supplement 1D). Flow cytometry analysis of CRISPRi cells showed that HMGCR KD increased proteasome activity, consistent with the reported effect of statin treatment (Figure 6–figure supplement 1E) (Van Der Kant et al., 2019). Nevertheless, in conditions that strongly modulated SM or CE levels (KD of MYLIP, PSAP, or LDLR), we did not see significant increase or decrease of proteasome activity (Figure 6H). In contrast, KD of IDI1, an enzyme downstream of HMGCR along the mevalonate pathway (but not FDFT1), strongly increased proteasome activity (Figure 6–figure supplement 1E). We conclude that the proteasome modulating effect of statin likely arises from changes in other metabolites along the mevalonate pathway. We summarize these results from testing the various mechanistic models in Figure 7.

Mechanistic models linking pathways identified in CRISPRi screens and lysosomal proteomics to accrual of lysosomal protein aggregates.

Discussion

Aging is associated with decreases in cellular proteostasis capacity and the accumulation of protein aggregates in diverse tissues. Despite extensive studies, factors modifying cellular proteostasis state, which has the potential to affect the state of many endogenous proteins, remain incompletely characterized. In this study, we performed genome-wide CRISPRi screens using a fluorescent dye sensitive to endogenous protein aggregation and uncovered links between lipid homeostasis and proteostasis. We discovered and characterized how changes in the uptake and metabolism of lipids, particularly SMs and CEs, lead to accumulation or depletion of these lipids and alter homeostasis of endogenous proteins.

SM and CE dysregulation has been implicated in various age-associated diseases. Intracellular sphingolipid accumulation has long been associated with lysosomal storage diseases, which share many features with age-related neurodegenerative diseases both at the level of physiology and cellular mechanism (Lloyd-Evans and Haslett, 2016). Elevated extracellular LDL cholesterol is associated with atherosclerotic cardiovascular disease (Mortensen and Nordestgaard, 2020), where modifications, including oxidation, and misfolding of apoB-100 can trigger the formation of Thioflavin T-positive fibrils of aggregated LDL and stimulate uptake by macrophages and catabolism within their lysosomes (Khoo et al., 1988; Parasassi et al., 2008). Consistent with our observations, MYLIP deficiency in mice leads to upregulated LDLR expression and increased cellular uptake of lipoprotein ApoE in brain tissues, but also increased uptake and clearance of extracellular aggregated Aβ peptides (Choi et al., 2015). These data affirm biological connections between lipoproteins and amyloidogenesis.

What are the molecular mechanisms that impair proteostasis in response to lipid perturbation? Accumulation of sphingolipid or cholesterol due to either direct enzyme deficiency or impaired lipid transport was previously linked to reduced cellular proteostasis via impairment of lysosomal function, manifested variously as reduced lysosomal proteolytic capacity, LMP, lipid peroxidation, and aberrant mTORC1 signaling (Kirkegaard et al., 2010; Gabandé-Rodríguez et al., 2014; Platt et al., 2012; Davis et al., 2021). Surprisingly, the lipid perturbations associated with ProteoStat-positive aggregates did not impinge on general features of lysosomal health. More recently, Tian and colleagues discovered in iPSC-induced neurons that PSAP KD severely impairs GSL catabolism, leading to the accumulation of these lipids and, in turn, lipofuscin formation, lipid peroxidation, and eventual cell death by ferroptosis (Tian et al., 2021). Lipofuscins are lysosomal deposits of heterogeneous nature, consisting mainly of oxidized proteins, lipids, metal ions, and sugar residues (Moreno-García et al., 2018). These deposits are largely seen in post-mitotic and senescent cells, and their accumulation is observed in aged tissues but also associated with many age-related degenerative diseases (Moreno-García et al., 2018). PSAP KD did not induce lipofuscin in rapidly dividing K562 cells (Figure 6–figure supplement 1F), likely due to continuous dilution of damaged biomolecules by rapid cell divisions before their levels can reach a threshold to form crosslinked and undegradable granules. Despite no visible accumulation of lipofuscin, we nonetheless observed an increase in lipid peroxidation and concomitant increase in lysosomal content, similarly to what’s reported in iPSC-derived neurons (Tian et al., 2021), suggesting that granular lipofuscin formation may be dispensable for ROS generation. Furthermore, antioxidants could uncouple lipid peroxidation from lysosomal protein aggregation induced by lipid dysregulation, thus ruling out a causal relationship.

ESCRT impairment was previously linked to LMP and ESCRT-III knockdown was shown to promote the escape of exogenous tau seeds from endolysosomes (Chen et al., 2019). Here we demonstrate that cholesterol and sphingolipid dysregulation can disrupt ESCRT function at the lysosome, leading to accumulation of detergent insoluble components that resist the action of VPS4A. Lipid composition might directly interfere with ESCRT function and intraluminal vesicle formation. While ESCRT-III polymerization can curve membranes and drive vesicle formation (McCullough et al., 2015), membrane curvature is also influenced by the shape and structure of lipid species present in the bilayer. Remarkably, enzymatic conversion of SM into ceramide (promoting neutral and negative membrane curvature, respectively) is sufficient to trigger spontaneous vesicle budding into the lumen of giant unilamellar vesicles in the absence of ESCRT (Trajkovic et al., 2008). Accordingly, formation of intraluminal vesicles in a mouse cell line is impaired when SM hydrolysis is inhibited (Trajkovic et al., 2008). Furthermore, SM hydrolysis promotes lysosomal repair in HeLa cells regardless of ESCRT status, and sphingomyelinase inhibition is detrimental to the process (Niekamp et al., 2022). Hence, an overabundance of SM may hinder the coalescence of ceramide-rich domains and prevent negative membrane curvature, thereby thwarting ESCRT resolution and vesicle budding. We show that this can in turn impair proteostasis and lead to aggregate accumulation, presumably as misfolded proteins are recruited to endolysosomal surfaces through microautophagy but fail to be degraded in the lumen.

Taken together, the data presented here demonstrate that, despite the general resilience of the proteostasis network, it is sensitive to pleiotropic changes in lysosomal lipid metabolism. Proteostasis is thus impacted by both changes in cell-intrinsic function and systemic lipid metabolism that can influence circulating lipids. This axis is of particular interest in the context of aging physiology in its ability to link processes that operate at the organismal and tissue levels to cellular proteostasis.

Materials and methods

Key Resources Table

Cell line construction and screen procedure

CRISPRi cell line was obtained by transducing K562 cells with a lentiviral vector expressing the nuclease-deactivated Cas9 fused with transcriptional repressor KRAB (UCOE-SFFV-KRAB-dCas9), followed by antibiotic selection. Monoclonal cell lines were obtained by limiting dilution, and screened functionally for effectiveness at silencing test genes when treated with sgRNAs. Cells were maintained in suspension at a density between 125K to 1M cells/mL in RPMI complete media (with 10% FBS and 1x GlutaMax). 500 million stable CRISPRi cells were transduced with a sgRNA lentivirus pool (hCRISPRi_v2.1_top5 library, which includes 102,640 gene-targeting and 4,590 non-targeting sgRNAs) at an MOI of 0.3 for single integrants. After 6 days in culture, including antibiotic selection for cells carrying the sgRNAs with 2ug/mL puromycin, these cells were stained with fixable blue dead cell stain (Molecular Probes), then fixed, permeabilized and stained with ProteoStat Aggresome detection kit (Enzo) according to manufacturer’s protocol. Stained cells were sorted by FACS (BD FACSAria) into top and bottom quartiles (ProteoStat-high and ProteoStat-low, respectively), using gates that excluded dead cells and corrected for correlation between cell size and staining intensity. Sorted cell populations were pelleted, from which genomic DNA was extracted with phenol:chloroform:isoamyl alcohol and ethanol precipitation. sgRNA-containing sequences were amplified from the genomic DNA with two rounds of PCR, introducing multiplexing barcodes and sequencing adaptors in the process. Amplicons were then purified and sequenced with the HiSeq4000 platform (Illumina). Sequencing data were analyzed using the MAGeCK pipeline (Li et al., 2014). GSEA was performed using the GSEAPY package (Subramanian et al., 2005; Mootha et al., 2003) with the prerank method, using the following score that takes into similarly weighted accounts both the effect size and confidence level for each gene: where nlFDRgene = log10(1/FDRgene), and D = 1 or – 1 such that the overall sign matches that of LFCgene.

Validation of hit genes by batch-retesting a reduced pool of CRISPRi guides

A subset of sgRNAs were selected and cloned into a reduced-scale library that consisted of 572 sgRNAs targeting 286 genes (87 enriched in ProteoStat-high, 99 enriched in ProteoStat-low, and 100 unenriched but abundantly expressed in K562 cells) and 300 non-targeting sgRNAs. sgRNA transduction in K562 CRISPRi cells, ProteoStat staining, FACS, library preparation and sequencing were performed similarly to the genome-wide experiment, except at a reduced scale (20 million cells transduced with sgRNA lentivirus pool, and >2.8 million cells each of ProteoStat-high and ProteoStat-low populations were obtained by FACS).

ProteoStat assays on cells with individual perturbations

K562 cells in multi-well plates were treated with individual perturbations and harvested for ProteoStat staining. To control for cell density-dependent variation in staining, harvested cells were spiked with untreated cells carrying appropriate fluorescent markers (e.g. TagBFP, mCerulean, or unlabeled, to be distinguished from treated cells) before fixation, permeabilization, and staining. ProteoStat fluorescence was quantified by flow cytometry (Fortessa, BD) with excitation at 488 nm and bandpass emission at 695/40 nm. ProteoStat intensity for each sample was first normalized to the internal spiked in (untreated) control, then to the corresponding negative control sample in the experiment (e.g. Non-targeting sgRNA). Individual perturbations were achieved by CRISPRi (sgRNA lentiviruses) or pharmacological inhibitors -- MG132, Bafilomycin A1, VER155008, Ganetespib, 17-DMAG, MBCD, Bortezomib, LLOMe, and Cumene Hydroperoxide. Lipid deprivation was performed by substituting regular FBS with delipidated FBS (Lipoprotein depleted FBS).

Immunofluorescence and cell staining by fluorescent probes

Cholesterol quantification was performed by formaldehyde fixing and staining cells with Filipin III according to manufacturer’s instructions.

Immunofluorescence assays were performed with standard protocol. Briefly, cells were washed with PBS and fixed with 4% formaldehyde for 30 minutes at room temperature, then washed with PBS. Cells were permeabilized in blocking buffer (PBS +0.5% Triton-X 100 +5% goat serum) for 1 hour at room temperature, followed by incubating with blocking buffer + primary antibody at 4°C overnight. Cells were then washed 3 times with PBS-T (PBS + 0.05% v/v tween20) at room temperature and incubated with blocking buffer + secondary antibody at room temperature for 1 hour. Cells were washed twice more with PBS-T and once with PBS before imaging. If co-staining with ProteoStat, the dye was included at 1:10000 during both primary and secondary antibody incubation. List of antibodies used: mouse-anti-LAMP1, mouse-anti-mono/polyubiquitinated conjugates, and Alexa Fluor Plus 647-conjugated goat-anti-mouse. Lipid peroxidation was quantified by incubating cells with Liperfluo at 5uM for 1 hour at 37°C. Cells were resuspended in cold PBS before analysis with flow cytometry (Fortessa, BD). Antioxidants, when included, consist of the following components at the corresponding final concentration: 5 ug/mL catalase, 2 ug/mL glutathione, 15 U/mL superoxide dismutase, 2 ug/mL DL-alpha tocopherol, and 2 ug/mL DL-alpha tocopherol acetate.

Lysosomal protease activity was monitored with DQ-Red BSA. K562 CRISPRi cells carrying sgRNAs targeting corresponding genes (7 days after sgRNA introduction) were pelleted and resuspended in prewarmed trafficking media (RPMI + 1% serum + 1% GlutaMax + 10 nM HEPES) containing 10 ug/mL DQ-Red BSA and returned to incubator at 37°C. After incubation for 1 hour, cells were washed once with cold PBS, resuspended in cold PBS, and analyzed by flow cytometry (Fortessa, BD) with excitation at 561 nm and bandpass emission at 610/20 nm.

Lysosomal content was estimated by incubating cells under various genetic perturbations (7 days after sgRNA transduction) with LysoTracker Red DND-99 at 100 nM. After incubation at 37°C for 1 hour, cells were washed once with cold PBS, resuspended in cold PBS, and analyzed by flow cytometry (Fortessa, BD) with excitation at 561 nm and bandpass emission at 610/20 nm.

Proteasome activity was assessed with flow cytometry by incubating cells with genetic (7 days) or chemical (1 hour) perturbations with Me4BoVS. After incubation at 37°C for 2 hours, cells were washed once with cold PBS, resuspended in cold PBS, and analyzed by flow cytometry (Fortessa, BD) with excitation at 488 nm and bandpass emission at 525/50 nm. For in-gel fluorescence analysis, cells treated with chemical inhibitors (1-20 hours) were incubated with Me4BoVS for 1 hour at 37°C, washed once with cold PBS and pelleted. Cells were lysed with cold NP40 lysis buffer (50 mM Tris-HCl pH7.4, 150 mM NaCl, 1% NP40, filtered) for 30 minutes at 4°C and pelleted at 14,000g for 3 minutes to remove debris. Samples were normalized by protein concentration (BCA assay) and run on a 4-20% Mini-PROTEAN TGX gel (Bio-Rad) and imaged directly on ChemiDoc with excitation at 460-490 nm and emission at 518-546 nm.

Lipidomics experiment and data analysis

K562 cells under various CRISPRi or lipid perturbations were maintained, in triplicate, for 7 days. Cells were pelleted, resuspended in fresh medium, and returned to culture on Day 6 -- one day before up to ∼ 1 million cells per sample were harvested. Briefly, cells were pelleted and washed twice with PBS at 4°C, then lysed using 800 uL of ice-cold MeOH:H2O (1:1,v/v) containing 3% by volume of lipid standards - Avanti 330707 SPLASH® LIPIDOMIX® Mass Spec Standard, which includes 18:1(d7) Chol Ester, d18:1-18:1(d9) SM, and Cholesterol (d7), among other lipid classes) at -20°C, and transferred to a glass vial. Following the methyl tertiary-butyl ether (MTBE)-liquid-liquid extraction (LLE) adapted from the Matyash protocol (Matyash et al., 2008), 800 uL of 100% MTBE was added into each sample, vortexed, and incubated on ice for 10 minutes to form phase separation. Next, after centrifuging samples at 3000g at 4°C for 5 minutes, 500 uL of the upper organic phase was collected into a clean glass vial, and the lower phase was re-extracted by the addition of 600 uL of 100% MTBE as described above. The organic phases were combined, dried under nitrogen, and resuspended in 150 uL of HPLC graded-ButOH/MeOH/H2O (2:2:1,v/v/v) with internal standards including: 5 ug/mL PE(18:0/18:0)-D70, PA(14:0/14:0)-D54, and 25 ug/mL LPC (16:0)-D31. Samples were analyzed using Thermo Scientific Vanquish UHPLC coupled with Q Exactive Plus Mass Spectrometer with 7.5 uL injection volume.

Analysis via separate injections in negative and positive mode ionization were done on a Vanquish UPLC coupled to Q-Exactive Plus mass spectrometers (Thermo Scientific). Lipids were separated on a Thermo Accucore C30 (250 x 2.1 mm, 2.6 μm polymer particles). Mobile phase A was 20 mM ammonium formate in MeCN:water (60:40,v/v); mobile phase B was 20 mM ammonium formate in isopropanol:MeCN (90:10, v/v) and a flow rate of 0.2 mL/min was used. The column was equilibrated for 7 min in 30% B prior to injection; followed by a gradient of 30-43%B from 0 to 7 min, 43-65% B from 7 to 12 min, 65-70% B from 12 to 30 min, 88% B at 31 min, 88-95% B at 51 min, 100% B at 53 min and 55 min, and 30%B from 55 to 60 min.

The source parameters for the MS were as follows: Sheath gas: 40; Aux gas:15; Sweep gas: 1; spray voltage 3.1 kV, the capillary temperature of 275°C, and S-lens RF level of 50. The mass spectrometer was operated in data-dependent top-8 MS2 mode, with 14000 resolution setting and AGC of 3e6 for MS1, and 17,500 resolution setting and AGC target of 3e6 for MS2. Stepped, normalized collision energies of 20, 30, 40 were used.

Data were analyzed and visualized using the open source software MAVEN peak analysis program (https://github.com/eugenemel/maven) (Seitzer et al., 2022). Compounds were identified with the criteria of a precursor ion tolerance of 10 ppm and a product ion tolerance of 20 ppm, comparing fragmentation and retention time to in-house generated in-silico libraries for lipidomics (https://github.com/calico/CalicoLipidLibrary).

PSAP variant series and rescue experiments

A series of constructs expressing various PSAP variants were built into lentiviral vectors under a UBC promoter. The canonical sequence mRNA variant 1 (NM_002778.4) was designated as wild-type. The full-length, wild-type sequence was modified using the following known disease-associated mutations to generate deficient variants: SapA-deficiency (p.V70del), SapB-deficiency (p.N215H_C241S), SapC-deficiency (p.L349P_C388F). The SapD-deficiency variant was created by truncating the entire C-terminal starting at the first SapD residue (p.405_524del). These mutations are procured from OMIM (https://omim.org/entry/176801) and other sources in literature (Spiegel et al., 2005; Schnabel et al., 1991). PSAP KD rescue experiments were performed by co-transducing K562 CRISPRi cells with lentiviruses expressing either a PSAP-targeting or non-targeting sgRNA, and in parallel one of the various PSAP variants. Transduced cells were selected with 2 ug/mL puromycin and 500 ug/mL geneticin for doubly transduced cells. Cells were harvested 6 days after transduction, spiked with untreated cells as internal control, then stained with ProteoStat and analyzed with flow cytometry as in other experiments.

Lyso-IP, silver stain, differential extraction

Lyso-IP cells were obtained by transducing K562 CRISPRi cells with a lentiviral vector expressing a lysosomal membrane protein fused with an epitope and fluorescent protein (UCOE-SFFV-TMEM192-mRFP1-3xHA-2A-Neo), followed by antibiotic selection. Cells were transduced in triplicates with corresponding sgRNA lentivirus and expanded in culture for 6 days, including antibiotic selection with 2ug/mL puromycin for cells carrying the sgRNAs. Lysosomes were harvested in a protocol partly adapted from the Sabatini Lab (Abu-Remaileh et al., 2017). Briefly, up to 40 million cells per sample were harvested, processing quickly on ice or at 4°C in batches of 4-6 samples. Cells were washed twice with chilled 10-25 mL PBS by pelleting at 1,000g for 2 minutes each, followed by resuspension in 1 mL KPBS (136 mM KCl, 10 mM KH2PO4, adjusted to pH 7.25 with KOH) and a final centrifugation in an 1.5 mL low-retention Eppendorf tube. Cell pellets were resuspended in 950 uL KPBS + inhibitors (Halt Protease and Phosphatase Inhibitor Cocktail (from 100x) + 0.5 mM PMSF). One at a time, cell suspensions were passed 20 times through a cell homogenizer (Isobiotec) with a 8 um clearance. Cell homogenates were clarified by pelleting at 1,000g for 2 minutes to remove cell debris and nuclei. Clarified supernatants were incubated with 100 uL of KPBS-prewashed anti-HA magnetic beads on rotation for 3 minutes and put on a DynaMag magnet rack. Lysosomes bound on beads were washed thrice with cold 1 mL KPBS directly on the magnet. After final buffer removal, washed beads were flash-frozen and stored at -80°C until lysis, when they were thawed briefly on ice.

Bound lysosomes were lysed by incubating in up to 150 uL (scaled with cell number) 1x RIPA complete buffer (RIPA (from Abcam 10x) + 2x Halt Protease and Phosphatase Inhibitor Cocktail (from 100x) + 0.5 mM PMSF + Benzonase (from 1000x) + 2 mM MgCl2) on rotation for 30 minutes. Beads were then removed with magnets and then again by centrifuging at 1,000g for 2 minutes. An aliquot of this Lysosomal Total fraction was set aside, and the rest is pelleted at 21,000g for 30 minutes to separate into the Lysosomal Soluble fraction (supernatant) and Lysosomal Insoluble fraction (pellet). The RIPA-insoluble pellets were washed 1x with 100 uL 1x RIPA complete buffer with brief vortexing (10 seconds), and pelleted again at 21,000g for 30 minutes. Final pellets were resuspended with up to 80 uL (scaled with cell number) 1x Urea/SDS complete buffer (8M Urea, 2% SDS, 50 mM DTT, 50 mM Tris-HCl, 1x Halt Protease and Phosphatase Inhibitor Cocktail (from 100x), pH 7.4), and bath-sonicated at low power for 10 minutes. All fractions were flash-frozen and stored at -80°C until further analysis.

For polyacrylamide gel and silver stain analysis, 30 uL per Lysosomal Insoluble fractions were loaded onto each lane of a 4-20% Mini-PROTEAN TGX gel (Bio-Rad) and stained with Pierce Silver Stain Kit, and imaged on the Odyssey CLx Imager (LI-COR) according to manufacturer’s protocols.

Protein Extraction and TMT Labeling for Proteomic Analysis

Lysosomal Total (in 1x RIPA complete buffer) and Lysosomal Insoluble (in 1x Urea/SDS complete buffer) samples (see above section for protocol and buffer compositions) were prepped using a combined manual version of our in-house automated multiplexed proteome profiling platform (AutoMP3) protocol (Gaun et al., 2021) and the single-pot, solid-phase-enhanced sample-preparation protocol (Hughes et al., 2019) with some minor modifications. In brief, 30 µL of the samples were diluted with 50 µL PBS. Samples were heated to 95°C for 5 minutes, briefly cooled and then treated with benzonase for 30 minutes at 37°C followed by sonication for 5 minutes. Samples were reduced with 5 mM DTT at 37°C for 30 minutes, alkylated with 15mM iodoacetamide for 20 minutes in the dark and then quenched with 5 mM DTT for 15 min at room temperature in the dark. Protein cleanup was performed using a 1:1 mixture of E3:E7 Sera-Mag Carboxylate-Modified Magnetic Particles (Cytiva Life Sciences, Marlborough, MA, USA). In brief, a bead stock was made and first washed with water 4 times, and then diluted to (20 µg solids / µL) and 100 µL beads were taken out into a new tube, placed on a magnet and the water removed followed by the addition of 100 µL of the sample to the beads and then mixed thoroughly off the magnet. The mixture was then immediately brought to 75% acetonitrile by addition of 300 µL of 100% acetonitrile and then incubated 18 min at room temp, allowing beads to aggregate with proteins. Tubes were then immobilized on the magnet and the supernatant removed. The beads were washed twice with 70% ethanol, and once more with 100% acetonitrile before they were brought into suspension with 92 µL digestion buffer (50mM EPPS pH 8.5, 10mM CaCl2) and Trypsin/LysC (Promega) was added at a ratio of 1:25 (enzyme: substrate) and incubated for 16 hours at 37°C with shaking (800 rpm). The next day vials were sonicated for 5 minutes and tubes were then immobilized on the magnet and the digest transferred to new vials. For TMT-labeling 40 µL of the digest was then mixed with 12 µL TMTpro-reagent (0.2 mg/ml, in acetonitrile) followed by incubation at 25°C for 1hr. All samples were quenched with 11 µL 5% hydroxylamine solution before they were combined. The combined samples were mixed and split into multiple peptide cleanup reactions in order to be able to reach the high % of needed organic solutions for peptide binding to beads. In brief, 50 µL prewashed beads were mixed with 50 µL of the combined sample and then the solution was immediately brought to 95% isopropanol by addition of 100% isopropanol. These samples were incubated at room temperature for 18 minutes to enable peptide binding to the beads. Tubes were then placed on the magnet for 2 min and the supernatant removed followed by two rounds of washes with 95% isopropanol, and then once with 100% acetonitrile. The peptides were then eluted in two rounds of 40 uL 5% acetonitrile followed by shaking at 1500 rpm for 1 minute. The cleaned up TMT-labeled samples were dried down in a speedvac (Labconco) and then resuspended in 5% formic acid, 5% acetonitrile (targeting 0.25 µg/µL) and 4 µL was injected for analysis per run.

Mass Spectrometry Proteomic Analysis

All data were obtained on an Orbitrap Eclipse mass spectrometer with FAIMS (Field Asymmetric Ion Mobility Spectrometry) Pro Interface. The mass spectrometer was coupled to an UltiMate 3000 HPLC operating in DDA mode (Thermo Fisher Scientific). Peptides were separated on an Aurora Series emitter column (25 cm × 75 µm i.d., 1.6 µm, 120 Å pore size, C18; IonOpticks) using a 165 min gradient from 8 to 30% acetonitrile in 0.125% formic acid. FAIMS was enabled during data acquisition with compensation voltages set as −40, −50, −60 and -70 and each voltage had a cycle time of 1.25 s. A high-resolution MS1 scan was performed in the Orbitrap at a 120,000 resolving power, m/z range 400–1600, RF lens 30%, standard automatic gain control (AGC) target, and “Auto” maximum injection time and data collected in profile mode. For MS2 collision-induced dissociation (CID) was used and the fragments were analyzed in the ion trap. A 0.7 m/z isolation window was used and the normalized collision energy for CID was 35% with a CID activation time of 10 ms at 0.25 activation Q and AGC 1 × 104 with 35 ms maximum injection time and data collected in centroid mode. Real-time search (RTS) (Schweppe et al., 2020) with the Uniprot database UP000005640 was used. Depending on the RTS, the ions were then analyzed with SPS MS3. For the RTS, the minimum Xcorr was set to 1, the minimum dCn was set to 0.1, and the maximum missed cleavages allowed was set to two. MS3 analysis was performed in the Orbitrap at a 50,000 resolving power, with HCD activation and 45% normalized collision energy, AGC 1 × 105, m/z range 100–500 m/z, and 200 ms maximum injection time and data collected in centroid mode. A maximum of 10 fragment ions from each MS2 spectrum were selected for MS3 analysis using SPS.

Computational Interpretation of Proteomic Data

An in-house software pipeline (version 3.12) was used to process all mass spectrometry data (Huttlin et al., 2010). Raw files were first converted to mzXML files using the MSConvert program (version 3.0.45) to generate peak lists from the RAW data files, and spectra were then assigned to peptides using the SEQUEST (version 28.12) algorithm (Eng et al., 1994). Spectra were queried against a “target-decoy” protein sequence database consisting of human proteins (using the reviewed Swiss-Prot portion of UP000005640 containing 20596 features and common contaminants) in forward and reversed decoys of the above (Elias and Gygi, 2007). The parent mass error tolerance was set to 20 ppm and the fragment mass error tolerance to 0.6 Da. Trypsin specificity was required allowing for up to two missed cleavages. Carbamidomethylation of cysteine (+57.02 Da), TMTpro-labeled N terminus and lysine (+304.20 Da) were set as static modifications. Methionine oxidation (+15.99 Da) were set as variable modifications. Following database searching, linear discriminant analysis was performed to filter peptide spectral matches to a 1% false discovery rate (FDR) (Huttlin et al., 2010). Following peptide filtering, non-unique peptides were assigned to proteins that comprised the largest number of matched redundant peptide sequences using the principle of Occam’s razor. The quantification of TMT reporter ions was performed by extracting the most intense ion at the predicted m/z value for each reporter ion (within a 0.003 m/z window) and isotopic purity correction of reporter quant values was applied. Known false positives (i.e., decoys and contaminants) were excluded from further analysis steps and peptide intensities and signal-to-noise ratios were exported for further analysis. In order to identify differentially expressed proteins between each condition the msTrawler v1 statistical software package (O’Brien et al., 2024) was used.

Lysosomal membrane permeability assay

Stable LMP reporter cells were generated by transducing K562 CRISPRi cells with a lentiviral vector expressing TagBFP-Gal3, followed by selection with 10 ug/mL blasticidin. To measure the effect of gene KDs on LMP, these reporter cells were transduced with corresponding sgRNA-expressing lentiviruses and selected with 2 ug/mL puromycin. 5 days after sgRNA transduction, cells were plated on poly-D-lysine-coated CellCarrier Ultra 96-well plates (Perkin Elmer) at 12K cells in 200 uL regular media per well. 2 days after plating, cells were treated with 60 uM LLOMe (or left untreated) and imaged on Opera Phenix imaging platform (Perkin Elmer) at 15-minute intervals for 4 hours, with excitation laser at 375 nm and bandpass emission filter at 435-550 nm. Images were analyzed with Harmony software (Perkin Elmer) by cell segmentation and spot detection. The number of Gal3 puncta in each field of view (6 fields per condition) is normalized by the total segmented cell area and scaled to match the average cell area, to correct for cell density and averaged across replicates. The maximal response for each sample over the 4-hour time course was reported.

pHLys quantification by pHLARE

Stable pHLys reporter cells were generated by transducing K562 CRISPRi cells with a lentiviral vector expressing pHLARE (a fusion protein consisting of prolactin signal sequence, superfolder GFP [sfGFP], rat LAMP1, and mCherry developed by Webb et al.) (Webb et al., 2021), and selected with 10 ug/mL blasticidin.

For validation, reporter cells were pelleted at 500 g for 5 minutes and resuspended in a buffer series at pHs ranging from 4 to 7.5 (140 mM KCl, 1 mM MgCl2, 1 mM CaCl2, 5 mM glucose and 10% citrate-phosphate-borate buffer at the target pH) (Carmody, 1961) in the presence of ionophores (10 uM nigericin, 2 uM monensin). Non-reporter cells were treated in parallel to correct for changes in background autofluorescence under various perturbations. After incubation at 37°C for 10 minutes, cells were analyzed with flow cytometry (Fortessa, BD), using excitation-bandpass emission settings of 488 nm-525/50 nm and 561 nm-610/20 nm for sfGFP and mCherry, respectively. To elucidate the effect of gene KDs on pHLys, CRISPRi cells with or without stable pHLys reporter were transduced with sgRNA-expressing lentiviruses, selected with 2 ug/mL puromycin, and analyzed similarly with flow cytometry, 7 days after sgRNA transduction.

Cytometry data was analyzed in FlowJo software (BD). Median background fluorescence in each channel in the cognate non-reporter controls was subtracted from the sfGFP and mCherry fluorescence intensities in the pHLys reporter samples. The ratio of adjusted sfGFP intensity vs adjusted mCherry intensity was reported.

Giant unilamellar vesicles (GUVs) and ProteoStat staining

GUVs with binary mixtures of saturated and unsaturated lipids were obtained by electroformation. Briefly, DOPC, DPPC, and free cholesterol were mixed in equal molar parts to a final concentration of 1mg/mL. TopFluor Cholesterol (Avanti Polar Lipids Inc.) was included at 1% of total lipids. The lipid mix was spread onto ITO-coated glass coverslips and dried overnight under vacuum. Dried lipids were hydrated with 300-400 ul of a 100 mM sucrose solution while applying a sinusoidal voltage (1.2 Vp-p and 10Hz) for one hour at 60°C. GUVs were harvested, diluted 20-fold in 200 uL of a 100 mM glucose solution supplemented with either no ProteoStat, or ProteoStat at a final dilution of 10,000x and 500x and imaged at 30°C in a spinning disk confocal fluorescence microscope (Nikon), using excitation/emission at 488 nm/525±18 nm (TopFluor) or 561 nm/605±35 nm (ProteoStat). Multi-channel single-plane images were acquired using NIS-Elements software and analyzed with ImageJ/Fiji (Schindelin et al., 2012).

Transmission Electron Microscopy

K562 CRISPRi cells carrying either non-targeting or PSAP-targeting sgRNA (6 days after sgRNA transduction) were fixed by directly mixing cells in suspension 1:1 with prewarmed 1x fixative (2% glutaraldehyde and 4% paraformaldehyde in 0.1 M sodium cacodylate buffer, pH7.4), followed immediately by gentle pelleting at 100 g for 10 minutes. Cell pellets were resuspended gently with fresh 1x fixative, and incubated at room temperature for 1 hour, followed by washing once with 2 mL 0.1 M sodium cacodylate buffer at room temperature and then transferring to 0.5 mL 0.1 M sodium cacodylate buffer, each time pelleting at 100 g for 10 minutes.

Fixed cells were stored at 4°C until they were enrobed in gelatin and post-fixed with 1% osmium tetroxide, washed with ddH2O and stained with 1% uranyl acetate. Stained samples were dehydrated in a graded series of ethanol, infiltrated and embedded in epon resin. Samples were then trimmed and sectioned with an ultramicrotome and imaged on a JEM-1400 transmission electron microscope (JEOL).

Materials availability

All new CRISPRi and reporter cell lines, and the pooled sgRNA plasmid library are available from the corresponding author.